---
title: "Gene Age and Stress Adaptation: A Comparative Transcriptomic Analysis of Phosphate Starvation in Budding Yeast"
author: "Joshua Ayelazuno"
date: "`r Sys.Date()`"
output: 
  pdf_document: default
  html_document: default
  html_notebook: default
---
Goal. Quantify how phosphate-limitation (-Pi) transcriptional responses are distributed across orthology classes (single-core, multi-core, non-core) in four yeasts—S. cerevisiae (Scer), C. glabrata (Cgla), C. albicans (Calb), and K. lactis (Klac)—and test whether specific classes are enriched for differentially expressed genes (DEGs).

Biological question. Are stress-responsive genes preferentially drawn from duplicated (multi-copy) or lineage-specific (non-core) gene sets, as suggested by Doughty et al., or from conserved single-copy (single-core) genes? Do patterns differ by species?

######################################
Load raw counts (Salmon)
Scer, Cgla, Calb, Klac each import the species-specific salmon.merged.gene_counts.tsv tables.
These are the per-gene count matrices that DESeq2 will ingest.

Set output locations
*_results paths point to where species-specific DE/summary CSVs will be written.
*_plots paths point to where figures for each species will be saved.

Using here() keeps paths project-relative for reproducibility.

Load orthology mapping
orthogroups reads the consolidated OrthoFinder table (species_orthology.tsv).
This file provides per-species gene lists per orthogroup and is later used to assign core status (single-core, multi-core, non-core) to each gene.


```{r loadrequired-packages, warning=FALSE, message=FALSE}
#if (!requireNamespace("BiocManager", quietly = TRUE))
    #install.packages("BiocManager")
#BiocManager::install("GenomicFeatures")
#BiocManager::install("cqn") #optional for cqn normalisation
#BiocManager::install("DESeq2")## 
#BiocManager::install("tximport")
#BiocManager::install("apeglm")
#BiocManager::install("ggplot2")
#BiocManager::install("dplyr")
#BiocManager::install("org.Sc.sgd.db")
#BiocManager::install("clusterProfiler")
#BiocManager::install("tidyverse")
#BiocManager::install("here")
#BiocManager::install("data.table")

############ install topGo for functional analysis of DESeq genes 
#if (!requireNamespace("BiocManager", quietly=TRUE))
     #install.packages("BiocManager")
 #BiocManager::install("topGO")

#### load required packages 
#library(GenomicFeatures)
library(DESeq2)
#library(tximport)
library(biomaRt)
library(apeglm)
library(ggplot2)
library(dplyr)
#library(org.Sc.sgd.db)
#library(clusterProfiler)
library(tidyverse)
library(readxl)
library(here)
library(data.table)
###### Load species raw count matrices
####### S. cerevisiae
Scer <- read_tsv(here("05.metadata", "Scer", "salmon.merged.gene_counts.tsv"), show_col_types = FALSE)

####### C. glabrata 
Cgla <- read_tsv(here("05.metadata", "Cgla", "salmon.merged.gene_counts.tsv"), show_col_types = FALSE)

####### C. albicans 
Calb <- read_tsv(here("05.metadata", "Calb", "salmon.merged.gene_counts.tsv"), show_col_types = FALSE)

####### K. lactis 
Klac <- read_tsv(here("05.metadata", "Klac", "salmon.merged.gene_counts.tsv"), show_col_types = FALSE)

##### Create file paths for output
Scer_results <- here("06.results", "Scer")
Cgla_results <- here("06.results", "Cgla")
Calb_results <- here("06.results", "Calb")
Klac_results <- here("06.results", "Klac")
combined <- here("06.results", "combined")

###### Directories for analysis/plots
Scer_plots <- here("08.plots", "Scer")
Cgla_plots <- here("08.plots", "Cgla")
Calb_plots <- here("08.plots", "Calb")
Klac_plots <- here("08.plots", "Klac")


############ load orthogroup data file 
orthogroups <- read_tsv(here("02.annotationdata", "species_orthology.tsv"),  show_col_types = FALSE)

```

Differential gene expression analysis with DESeq2

Goal. Estimate gene-level differential expression between control (t0000) and treatment (t0060) for each species, collapsing technical replicates, and produce (i) a full results table for enrichment “universe” analyses and (ii) clean DEG sets (all / up / down) plus QC plots.

What this chunk does (re-usable for any species):
Select samples for the contrast.
Keep only columns matching t0000|t0060; coerce counts to integers; set gene_id as rownames.

Build sample metadata and handle replicates.
Derive condition from column names (untreated, treated).
Create sample IDs to group technical runs (e.g., 1–4) and set run to the original column names.

Construct DESeq2 object and collapse technical replicates.
DESeqDataSetFromMatrix(..., design = ~ condition) → model matrix.
collapseReplicates() sums counts within each biological sample across runs.
Re-apply the condition labels to the collapsed columns.

Fit the model and shrink effect sizes.
DESeq() fits NB GLMs.
results(..., contrast = c("condition","treated","untreated")) extracts treated vs control.
lfcShrink(..., type = "apeglm") returns stabilized log2FCs for reporting/plots.

Export the full universe for GO/topGO.
Convert to a data frame, clean gene_id (e.g., drop gene- prefix), and write <Species>_DE_topGo.csv (all tested genes, not just significant).

Call significant DEGs (and split by direction).
Apply thresholds padj < 0.01 and |log2FoldChange| > 1.
Split into resSig*_up (log2FC > 0) and resSig*_down (log2FC < 0).
Write three CSVs (all / up / down) and simple ID lists (*_UP_ids.txt, *_DOWN_ids.txt).

QC & exploratory visualizations.
rlog() → PCA plot to confirm replicate clustering and treatment separation.
Histograms of raw p-values and FDR to check multiple-testing behavior.
MA plot to inspect mean–FC trends.
Volcano plot with LFC/FDR cut lines to visualize significant genes.

Notes & conventions.
Contrast direction is treated vs untreated; positive log2FC means induced in treatment.\
Collapsing technical replicates improves variance estimates; biological replicates remain separate in the model.
Keep this workflow identical across species; only the input count matrix and output paths change.


```{r -Pi-Scer-t0060-Deseq-Analysis, warning=FALSE, message=FALSE}

# Automatically select all t0000 and t0060 replicates
rownames(Scer) <- Scer$gene_id
Scer0060min<- grep("t0000|t0060", colnames(Scer), value = TRUE)
sc0060min <- Scer[, Scer0060min]
sc0060min <- as.data.frame(lapply(sc0060min, as.integer))
rownames(sc0060min) <- rownames(Scer)


######################### DESeq

# Define sample metadata based on column names
conditions <- ifelse(grepl("t0000", Scer0060min), "untreated", "treated")
sc0060mins_coldata <- data.frame(
  row.names = Scer0060min,
  condition = factor(conditions, levels = c("untreated", "treated"))
)
############### Add biological replicate labels (to group technical replicates)
# This assigns replicate groups: 1,1,2,2,3,3,4,4
replicate_ids <- rep(paste0("sample", 1:4), each = 2)

sc0060mins_coldata$sample <- replicate_ids
sc0060mins_coldata$run <- rownames(sc0060mins_coldata)  # must match colnames of count matrix

########### Create DESeqDataSet and collapse technical replicates
# Create DESeqDataSet with full colData (including technical replicate info)
scdds <- DESeqDataSetFromMatrix(countData = sc0060min,
                              colData = sc0060mins_coldata,
                              design = ~ condition)

# Collapse technical replicates by summing counts
scdds_collapsed <- collapseReplicates(scdds,
                                    groupby = sc0060mins_coldata$sample,
                                    run = sc0060mins_coldata$run)

# Update condition info (after collapse, only one col per sample remains)
colData(scdds_collapsed)$condition <- colData(scdds)[!duplicated(sc0060mins_coldata$sample), "condition"]

# Run DESeq2 and shrink fold changes
scdds_collapsed <- DESeq(scdds_collapsed)
sc_res <- results(scdds_collapsed, contrast = c("condition", "treated", "untreated"))

##resultsNames(scdds_collapsed)
resLFCsc <- lfcShrink(scdds_collapsed,
                    coef = "condition_treated_vs_untreated", 
                    type = "apeglm")

### Full results (for topGO universe)
scer_topGo <- as.data.frame(resLFCsc) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::mutate(gene_id = sub("^gene-", "", gene_id))

write.csv(scer_topGo,
          file = file.path(Scer_results, "Scer_DE_topGo.csv"),
          row.names = FALSE)


########## Filter for Significant DE Genes; based on Doughty et al 2020

resSigsc<- subset(scer_topGo, padj < 0.01 & abs(log2FoldChange) > 1)

resSigsc_up <- subset(resSigsc, log2FoldChange > 0) ## upregulated 
resSigsc_down <- subset(resSigsc, log2FoldChange < 0) ##d downregulated 
# Write to CSV
write.csv(as.data.frame(resSigsc),
          file = file.path(Scer_results, "20250826Scer.LFCgt2.padj0.01.csv"),
          row.names = FALSE) ## all DEGs 

write.csv(as.data.frame(resSigsc_up),
          file = file.path(Scer_results, "20250826Scer.LFCup.padj0.01.csv"),
          row.names = FALSE) ## upregulated 

write.csv(as.data.frame(resSigsc_down),
          file = file.path(Scer_results, "20250826Scer.LFCdown.padj0.01.csv"),
          row.names = FALSE) ## downregulated 

######### write just gene IDs 
writeLines(resSigsc$gene_id[resSigsc$log2FoldChange >= 1],
           file.path(Scer_results, "Scer_UP_ids.txt"))
writeLines(resSigsc$gene_id[resSigsc$log2FoldChange <= -1],
           file.path(Scer_results, "Scer_DOWN_ids.txt"))

# Run rlog transformation
rld_sc <- rlog(scdds_collapsed, blind = TRUE)
rld_mat_sc <- assay(rld_sc)

# PCA calculation
pca_sc <- prcomp(t(rld_mat_sc))
percentVar <- round(100 * (pca_sc$sdev^2 / sum(pca_sc$sdev^2)))

# Extract sample info
sc_sample_info <- as.data.frame(colData(scdds_collapsed))
sc_sample_info$sample <- rownames(sc_sample_info)

# PCA data frame
df_pca <- data.frame(
  PC1 = pca_sc$x[, 1],
  PC2 = pca_sc$x[, 2],
  sample = rownames(pca_sc$x),
  group = sc_sample_info$condition
)

# PCA plot
ScerPCA1 <- ggplot(df_pca, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 6, alpha = 0.8) +
  labs(
    title = "PCA - Scer t0000 vs t0060",
    x = paste0("PC1 (", percentVar[1], "%)"),
    y = paste0("PC2 (", percentVar[2], "%)")
  ) +
  theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.title = element_blank(),
    legend.text = element_text(size = 14),
    legend.key.size = unit(1, "cm")
  ) +
  xlim(-25, 25) +
  ylim(-7, 7)

ScerPCA1
#ggsave(filename = file.path(Scer_plots, "PCA_Scer_t0000_vs_t0060.pdf"),
       #plot = ScerPCA1, width = 8, height = 6)

# p-value histogram
#pdf(file.path(Scer_plots, "Histogram_pvalue_Scer.pdf"))
hist(resLFCsc$pvalue,
     breaks = 50, col = "grey",
     main = "Distribution of Raw p-values",
     xlab = "p-values", ylab = "Frequency",
     cex.main = 1.5, cex.lab = 1.2)
#dev.off()

# Adjusted p-value histogram
#pdf(file.path(Scer_plots, "Histogram_padj_Scer.pdf"))
hist(resLFCsc$padj,
     breaks = 50, col = "grey",
     main = "Distribution of Adjusted p-values (FDR)",
     xlab = "FDR", ylab = "Frequency",
     cex.main = 1.5, cex.lab = 1.2)
#dev.off()

# MA plot
#pdf(file.path(Scer_plots, "MA_plot_Scer.pdf"))
plotMA(resLFCsc, main = "MA Plot - Scer t0000 vs t0060", cex = 0.5)
#dev.off()

# Volcano plot
lfc_threshold <- 1
pval_threshold <- 0.01

tab <- data.frame(
  logFC = resLFCsc$log2FoldChange,
  negLogPval = -log10(resLFCsc$padj)
)

sig <- with(tab, abs(logFC) > lfc_threshold & negLogPval > -log10(pval_threshold))
#pdf(file.path(Scer_plots, "Volcano_plot_Scer.pdf"))
plot(tab, pch = 16, cex = 0.5,
     xlab = expression(bold(log[2]~fold~change)),
     ylab = expression(bold(-log[10]~FDR)),
     main = "Volcano Plot - Scer t0000 vs t0060")
points(tab[sig, ], col = "red", pch = 16, cex = 0.6)
abline(h = -log10(pval_threshold), col = "green3", lty = 2)
abline(v = c(-lfc_threshold, lfc_threshold), col = "blue", lty = 2)
#dev.off()




```


```{r -Pi-Cgla-t0060min-Deseq-analysis, warning=FALSE, message=FALSE}
# Prepare count matrix
rownames(Cgla) <- Cgla$gene_id
Cgla0060min <- grep("t0000|t0060", colnames(Cgla), value = TRUE)
cg0060min <- Cgla[, Cgla0060min]
cg0060min <- as.data.frame(lapply(cg0060min, as.integer))
rownames(cg0060min) <- rownames(Cgla)

# Sample metadata
conditions_cg <- ifelse(grepl("t0000", Cgla0060min), "untreated", "treated")
cg_coldata <- data.frame(
  row.names = Cgla0060min,
  condition = factor(conditions_cg, levels = c("untreated", "treated"))
)
cg_coldata$sample <- rep(paste0("sample", 1:4), each = 2)
cg_coldata$run <- rownames(cg_coldata)

# Create DESeqDataSet and collapse replicates
cgdds <- DESeqDataSetFromMatrix(countData = cg0060min,
                                colData = cg_coldata,
                                design = ~ condition)

cgdds_collapsed <- collapseReplicates(cgdds,
                                      groupby = cg_coldata$sample,
                                      run = cg_coldata$run)

colData(cgdds_collapsed)$condition <- colData(cgdds)[!duplicated(cg_coldata$sample), "condition"]

# Run DESeq and shrink LFC
cgdds_collapsed <- DESeq(cgdds_collapsed)
resLFCcg <- lfcShrink(cgdds_collapsed,
                      coef = "condition_treated_vs_untreated",
                      type = "apeglm")

### Full results (for topGO universe)
cgla_topGo <- as.data.frame(resLFCcg) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::mutate(gene_id = sub("^gene-", "", gene_id))

write.csv(cgla_topGo,
          file = file.path(Cgla_results, "Cgla_DE_topGo.csv"),
          row.names = FALSE)


########## Filter for Significant DE Genes; based on Doughty et al 2020

resSigcg<- subset(cgla_topGo, padj < 0.01 & abs(log2FoldChange) > 1)

resSigcg_up <- subset(resSigcg, log2FoldChange > 0) ## upregulated 
resSigcg_down <- subset(resSigcg, log2FoldChange < 0) ##d downregulated 
# Write to CSV
write.csv(as.data.frame(resSigcg),
          file = file.path(Cgla_results, "20250826Cgla.LFCgt2.padj0.01.csv"),
          row.names = FALSE) ## all DEGs 

write.csv(as.data.frame(resSigcg_up),
          file = file.path(Cgla_results, "20250826Cgla.LFCup.padj0.01.csv"),
          row.names = FALSE) ## upregulated 

write.csv(as.data.frame(resSigcg_down),
          file = file.path(Cgla_results, "20250826Cgla.LFCdown.padj0.01.csv"),
          row.names = FALSE) ## downregulated 

######### write just gene IDs 
writeLines(resSigcg$gene_id[resSigcg$log2FoldChange >= 1],
           file.path(Cgla_results, "Cgla_UP_ids.txt"))
writeLines(resSigcg$gene_id[resSigcg$log2FoldChange <= -1],
           file.path(Cgla_results, "Cgla_DOWN_ids.txt"))

# rlog and PCA
rld_cg <- rlog(cgdds_collapsed, blind = TRUE)
rld_mat_cg <- assay(rld_cg)

# Recalculate PCA if needed
pca_cg <- prcomp(t(rld_mat_cg))
percentVar_cg <- round(100 * (pca_cg$sdev^2 / sum(pca_cg$sdev^2)))

# Extract sample info
cg_sample_info <- as.data.frame(colData(cgdds_collapsed))
cg_sample_info$sample <- rownames(cg_sample_info)

df_pca_cg <- data.frame(
  PC1 = pca_cg$x[, 1],
  PC2 = pca_cg$x[, 2],
  sample = rownames(pca_cg$x),
  group = cg_sample_info$condition
)

# Adjusted PCA plot: auto-scaled axes
CglaPCA1 <- ggplot(df_pca_cg, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 6, alpha = 0.8) +
  labs(
    title = "PCA - Cgla t0000 vs t0060",
    x = paste0("PC1 (", percentVar_cg[1], "%)"),
    y = paste0("PC2 (", percentVar_cg[2], "%)")
  ) +
  theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.title = element_blank(),
    legend.text = element_text(size = 14),
    legend.key.size = unit(1, "cm")
  )
CglaPCA1

# Save new plot
#ggsave(filename = file.path(Cgla_plots, "PCA_Cgla_t0000_vs_t0060_adjusted.pdf"),
       #plot = CglaPCA1, width = 8, height = 6)


# p-value histogram
#pdf(file.path(Cgla_plots, "Histogram_pvalue_Cgla.pdf"))
hist(resLFCcg$pvalue, breaks = 50, col = "grey",
     main = "Distribution of Raw p-values",
     xlab = "p-values", ylab = "Frequency",
     cex.main = 1.5, cex.lab = 1.2)
#dev.off()

# adjusted p-value histogram
#pdf(file.path(Cgla_plots, "Histogram_padj_Cgla.pdf"))
hist(resLFCcg$padj, breaks = 50, col = "grey",
     main = "Distribution of Adjusted p-values (FDR)",
     xlab = "FDR", ylab = "Frequency",
     cex.main = 1.5, cex.lab = 1.2)
#dev.off()

# MA plot
#pdf(file.path(Cgla_plots, "MA_plot_Cgla.pdf"))
plotMA(resLFCcg, main = "MA Plot - Cgla t0000 vs t0060", cex = 0.5)
#dev.off()

# Volcano plot
lfc_threshold <- 1
pval_threshold <- 0.01

tab_cg <- data.frame(
  logFC = resLFCcg$log2FoldChange,
  negLogPval = -log10(resLFCcg$padj)
)

sig_cg <- with(tab_cg, abs(logFC) > lfc_threshold & negLogPval > -log10(pval_threshold))

#pdf(file.path(Cgla_plots, "Volcano_plot_Cgla.pdf"))
plot(tab_cg, pch = 16, cex = 0.5,
     xlab = expression(bold(log[2]~fold~change)),
     ylab = expression(bold(-log[10]~FDR)),
     main = "Volcano Plot - Cgla t0000 vs t0060")
points(tab_cg[sig_cg, ], col = "red", pch = 16, cex = 0.6)
abline(h = -log10(pval_threshold), col = "green3", lty = 2)
abline(v = c(-lfc_threshold, lfc_threshold), col = "blue", lty = 2)
#dev.off()
```


```{r -Pi-Calb-t0060min-Deseq-Analysis, warning=FALSE, message=FALSE}
# Prepare count matrix
rownames(Calb) <- Calb$gene_id
Calb0060min <- grep("t0000|t0060", colnames(Calb), value = TRUE)
calb0060min <- Calb[, Calb0060min]
calb0060min <- as.data.frame(lapply(calb0060min, as.integer))
rownames(calb0060min) <- rownames(Calb)

# Sample metadata
conditions_calb <- ifelse(grepl("t0000", Calb0060min), "untreated", "treated")
calb_coldata <- data.frame(
  row.names = Calb0060min,
  condition = factor(conditions_calb, levels = c("untreated", "treated"))
)
calb_coldata$sample <- rep(paste0("sample", 1:4), each = 2)
calb_coldata$run <- rownames(calb_coldata)

# Create DESeqDataSet and collapse technical replicates
calbdds <- DESeqDataSetFromMatrix(countData = calb0060min,
                                  colData = calb_coldata,
                                  design = ~ condition)

calbdds_collapsed <- collapseReplicates(calbdds,
                                        groupby = calb_coldata$sample,
                                        run = calb_coldata$run)

colData(calbdds_collapsed)$condition <- colData(calbdds)[!duplicated(calb_coldata$sample), "condition"]

# Run DESeq and LFC shrinkage
calbdds_collapsed <- DESeq(calbdds_collapsed)
resLFCcalb <- lfcShrink(calbdds_collapsed,
                        coef = "condition_treated_vs_untreated",
                        type = "apeglm")

### Full results (for topGO universe)
calb_topGo <- as.data.frame(resLFCcalb) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::mutate(gene_id)

write.csv(calb_topGo,
          file = file.path(Calb_results, "Calb_DE_topGo.csv"),
          row.names = FALSE)


########## Filter for Significant DE Genes; based on Doughty et al 2020

resSigca<- subset(calb_topGo, padj < 0.01 & abs(log2FoldChange) > 1)

resSigca_up <- subset(resSigca, log2FoldChange > 0) ## upregulated 
resSigca_down <- subset(resSigca, log2FoldChange < 0) ##d downregulated 
# Write to CSV
write.csv(as.data.frame(resSigca),
          file = file.path(Calb_results, "20250826Calb.LFCgt2.padj0.01.csv"),
          row.names = FALSE) ## all DEGs 

write.csv(as.data.frame(resSigca_up),
          file = file.path(Calb_results, "20250826Calb.LFCup.padj0.01.csv"),
          row.names = FALSE) ## upregulated 

write.csv(as.data.frame(resSigca_down),
          file = file.path(Calb_results, "20250826Calb.LFCdown.padj0.01.csv"),
          row.names = FALSE) ## downregulated 

######### write just gene IDs 
writeLines(resSigca$gene_id[resSigca$log2FoldChange >= 1],
           file.path(Calb_results, "Calb_UP_ids.txt"))
writeLines(resSigca$gene_id[resSigca$log2FoldChange <= -1],
           file.path(Calb_results, "Calb_DOWN_ids.txt"))


# rlog and PCA
rld_calb <- rlog(calbdds_collapsed, blind = TRUE)
rld_mat_calb <- assay(rld_calb)

pca_calb <- prcomp(t(rld_mat_calb))
percentVar_calb <- round(100 * (pca_calb$sdev^2 / sum(pca_calb$sdev^2)))

calb_sample_info <- as.data.frame(colData(calbdds_collapsed))
calb_sample_info$sample <- rownames(calb_sample_info)

df_pca_calb <- data.frame(
  PC1 = pca_calb$x[, 1],
  PC2 = pca_calb$x[, 2],
  sample = rownames(pca_calb$x),
  group = calb_sample_info$condition
)

# Fixed PCA plot with autoscaling
CalbPCA1 <- ggplot(df_pca_calb, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 6, alpha = 0.8) +
  labs(
    title = "PCA - Calb t0000 vs t0060",
    x = paste0("PC1 (", percentVar_calb[1], "%)"),
    y = paste0("PC2 (", percentVar_calb[2], "%)")
  ) +
  theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.title = element_blank(),
    legend.text = element_text(size = 14),
    legend.key.size = unit(1, "cm")
  )
CalbPCA1

#ggsave(filename = file.path(Calb_plots, "PCA_Calb_t0000_vs_t0060.pdf"),
       #plot = CalbPCA1, width = 8, height = 6)

# p-value histogram
#pdf(file.path(Calb_plots, "Histogram_pvalue_Calb.pdf"))
hist(resLFCcalb$pvalue, breaks = 50, col = "grey",
     main = "Distribution of Raw p-values",
     xlab = "p-values", ylab = "Frequency",
     cex.main = 1.5, cex.lab = 1.2)
#dev.off()

# adjusted p-value histogram
#pdf(file.path(Calb_plots, "Histogram_padj_Calb.pdf"))
hist(resLFCcalb$padj, breaks = 50, col = "grey",
     main = "Distribution of Adjusted p-values (FDR)",
     xlab = "FDR", ylab = "Frequency",
     cex.main = 1.5, cex.lab = 1.2)
#dev.off()

# MA plot
#pdf(file.path(Calb_plots, "MA_plot_Calb.pdf"))
plotMA(resLFCcalb, main = "MA Plot - Calb t0000 vs t0060", cex = 0.5)
#dev.off()

# Volcano plot
lfc_threshold <- 1
pval_threshold <- 0.01

tab_calb <- data.frame(
  logFC = resLFCcalb$log2FoldChange,
  negLogPval = -log10(resLFCcalb$padj)
)

sig_calb <- with(tab_calb, abs(logFC) > lfc_threshold & negLogPval > -log10(pval_threshold))

#pdf(file.path(Calb_plots, "Volcano_plot_Calb.pdf"))
plot(tab_calb, pch = 16, cex = 0.5,
     xlab = expression(bold(log[2]~fold~change)),
     ylab = expression(bold(-log[10]~FDR)),
     main = "Volcano Plot - Calb t0000 vs t0060")
points(tab_calb[sig_calb, ], col = "red", pch = 16, cex = 0.6)
abline(h = -log10(pval_threshold), col = "green3", lty = 2)
abline(v = c(-lfc_threshold, lfc_threshold), col = "blue", lty = 2)
#dev.off()
```


```{r -Pi-Klac-t0060min-Deseq-Analysis, warning=FALSE, message=FALSE}
# Prepare count matrix
rownames(Klac) <- Klac$gene_id
Klac0060min <- grep("t0000|t0060", colnames(Klac), value = TRUE)
klac0060min <- Klac[, Klac0060min]
klac0060min <- as.data.frame(lapply(klac0060min, as.integer))
rownames(klac0060min) <- rownames(Klac)

# Sample metadata
conditions_klac <- ifelse(grepl("t0000", Klac0060min), "untreated", "treated")
klac_coldata <- data.frame(
  row.names = Klac0060min,
  condition = factor(conditions_klac, levels = c("untreated", "treated"))
)
klac_coldata$sample <- rep(paste0("sample", 1:4), each = 2)
klac_coldata$run <- rownames(klac_coldata)

# Create DESeqDataSet and collapse technical replicates
klacdds <- DESeqDataSetFromMatrix(countData = klac0060min,
                                  colData = klac_coldata,
                                  design = ~ condition)

klacdds_collapsed <- collapseReplicates(klacdds,
                                        groupby = klac_coldata$sample,
                                        run = klac_coldata$run)

colData(klacdds_collapsed)$condition <- colData(klacdds)[!duplicated(klac_coldata$sample), "condition"]

# Run DESeq and LFC shrinkage
klacdds_collapsed <- DESeq(klacdds_collapsed)
resLFCklac <- lfcShrink(klacdds_collapsed,
                        coef = "condition_treated_vs_untreated",
                        type = "apeglm")

### Full results (for topGO universe)
klac_topGo <- as.data.frame(resLFCklac) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::mutate(gene_id = sub("^gene-", "", gene_id))

write.csv(klac_topGo,
          file = file.path(Klac_results, "Klac_DE_topGo.csv"),
          row.names = FALSE)


########## Filter for Significant DE Genes; based on Doughty et al 2020

resSigkl<- subset(klac_topGo, padj < 0.01 & abs(log2FoldChange) > 1)

resSigkl_up <- subset(resSigkl, log2FoldChange > 0) ## upregulated 
resSigkl_down <- subset(resSigkl, log2FoldChange < 0) ##d downregulated 
# Write to CSV
write.csv(as.data.frame(resSigkl),
          file = file.path(Klac_results, "20250826Klac.LFCgt2.padj0.01.csv"),
          row.names = FALSE) ## all DEGs 

write.csv(as.data.frame(resSigkl_up),
          file = file.path(Klac_results, "20250826Klac.LFCup.padj0.01.csv"),
          row.names = FALSE) ## upregulated 

write.csv(as.data.frame(resSigkl_down),
          file = file.path(Klac_results, "20250826Klac.LFCdown.padj0.01.csv"),
          row.names = FALSE) ## downregulated 

######### write just gene IDs 
writeLines(resSigkl$gene_id[resSigkl$log2FoldChange >= 1],
           file.path(Klac_results, "Klac_UP_ids.txt"))
writeLines(resSigkl$gene_id[resSigkl$log2FoldChange <= -1],
           file.path(Klac_results, "Klac_DOWN_ids.txt"))


# rlog and PCA
rld_klac <- rlog(klacdds_collapsed, blind = TRUE)
rld_mat_klac <- assay(rld_klac)

pca_klac <- prcomp(t(rld_mat_klac))
percentVar_klac <- round(100 * (pca_klac$sdev^2 / sum(pca_klac$sdev^2)))

klac_sample_info <- as.data.frame(colData(klacdds_collapsed))
klac_sample_info$sample <- rownames(klac_sample_info)

df_pca_klac <- data.frame(
  PC1 = pca_klac$x[, 1],
  PC2 = pca_klac$x[, 2],
  sample = rownames(pca_klac$x),
  group = klac_sample_info$condition
)

# PCA plot (auto-scaled)
KlacPCA1 <- ggplot(df_pca_klac, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 6, alpha = 0.8) +
  labs(
    title = "PCA - Klac t0000 vs t0060",
    x = paste0("PC1 (", percentVar_klac[1], "%)"),
    y = paste0("PC2 (", percentVar_klac[2], "%)")
  ) +
  theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 22, face = "bold"),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.title = element_blank(),
    legend.text = element_text(size = 14),
    legend.key.size = unit(1, "cm")
  )
KlacPCA1
#ggsave(filename = file.path(Klac_plots, "PCA_Klac_t0000_vs_t0060.pdf"),
       #plot = KlacPCA1, width = 8, height = 6)

# Histograms
#pdf(file.path(Klac_plots, "Histogram_pvalue_Klac.pdf"))
hist(resLFCklac$pvalue, breaks = 50, col = "grey",
     main = "Distribution of Raw p-values",
     xlab = "p-values", ylab = "Frequency",
     cex.main = 1.5, cex.lab = 1.2)
#dev.off()

#pdf(file.path(Klac_plots, "Histogram_padj_Klac.pdf"))
hist(resLFCklac$padj, breaks = 50, col = "grey",
     main = "Distribution of Adjusted p-values (FDR)",
     xlab = "FDR", ylab = "Frequency",
     cex.main = 1.5, cex.lab = 1.2)
#dev.off()

# MA plot
#pdf(file.path(Klac_plots, "MA_plot_Klac.pdf"))
plotMA(resLFCklac, main = "MA Plot - Klac t0000 vs t0060", cex = 0.5)
#dev.off()

# Volcano plot
lfc_threshold <- 1
pval_threshold <- 0.01

tab_klac <- data.frame(
  logFC = resLFCklac$log2FoldChange,
  negLogPval = -log10(resLFCklac$padj)
)

sig_klac <- with(tab_klac, abs(logFC) > lfc_threshold & negLogPval > -log10(pval_threshold))

#pdf(file.path(Klac_plots, "Volcano_plot_Klac.pdf"))
plot(tab_klac, pch = 16, cex = 0.5,
     xlab = expression(bold(log[2]~fold~change)),
     ylab = expression(bold(-log[10]~FDR)),
     main = "Volcano Plot - Klac t0000 vs t0060")
points(tab_klac[sig_klac, ], col = "red", pch = 16, cex = 0.6)
abline(h = -log10(pval_threshold), col = "green3", lty = 2)
abline(v = c(-lfc_threshold, lfc_threshold), col = "blue", lty = 2)
#dev.off()
```
Goal.
Summarize the number of significant DE genes (UP vs DOWN) per species and sanity-check a conserved marker (PHO84 ortholog) at the sample level.
What this chunk does (applies to all species):

Count DEGs by direction.
Uses your already filtered DE tables (resSigsc, resSigcg, resSigca, resSigkl).

Defines UP as log2FC > 0, DOWN as log2FC < 0.
Assumes significance was called with padj < 0.01 and |log2FC| > 1. If you’re replicating Doughty’s Fig. 1c (UP-only), adjust accordingly.

Summarize and plot totals.
Builds deg_summary (UP, DOWN, Total) across species.
Converts to long format and makes a grouped bar chart (Upregulated vs Downregulated) per species.

Gene-level expression checks (PHO84 orthologs).
Uses plotCounts() on the collapsed DESeq2 objects (*dds_collapsed) to extract normalized counts for each PHO84 ortholog.
Plots log-scaled counts by condition with jittered points to show replicate spread.

Why this step matters.
The PHO84 plots verify that a canonical phosphate transporter behaves as expected in each species and that replicate variance looks reasonable.

```{r explore-DEGs-count-plotCount}
# Count significant DEGs
sc_up   <- sum(resSigsc$log2FoldChange > 0)
sc_down <- sum(resSigsc$log2FoldChange < 0)

cg_up   <- sum(resSigcg$log2FoldChange > 0)
cg_down <- sum(resSigcg$log2FoldChange < 0)

calb_up   <- sum(resSigca$log2FoldChange > 0)
calb_down <- sum(resSigca$log2FoldChange < 0)

klac_up   <- sum(resSigkl$log2FoldChange > 0)
klac_down <- sum(resSigkl$log2FoldChange < 0)

# Combine into a data frame
deg_summary <- data.frame(
  Species = c("Scer", "Cgla", "Calb", "Klac"),
  Upregulated = c(sc_up, cg_up, calb_up, klac_up),
  Downregulated = c(sc_down, cg_down, calb_down, klac_down)
)
deg_summary$Total <- deg_summary$Upregulated + deg_summary$Downregulated

print(deg_summary)

# Convert to long format
deg_long <- deg_summary |>
  pivot_longer(cols = c("Upregulated", "Downregulated"),
               names_to = "Direction", values_to = "Count")

# Bar plot
p.sig.Degs <- ggplot(deg_long, aes(x = Species, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_manual(values = c("Upregulated" = "tomato", "Downregulated" = "steelblue")) +
  labs(title = "Differentially Expressed Genes by Species",
       y = "Significant DEGs (|LFC| > 1)",
       x = "Species") +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    legend.title = element_blank()
  )
p.sig.Degs
#ggsave("/Users/jayelazuno/Desktop/EvoPhosphoOmics/plots/Sig.DEGs.species.pdf", 
       #plot = p.sig.Degs, width = 8, height = 6)

#########################plot counts for specific geness  
# Extract normalized counts for scPHO84 (YML123C)
scPHO84_counts <- plotCounts(scdds_collapsed,
                           gene = "gene-YML123C",
                           intgroup = "condition",
                           returnData = TRUE)

# Plot with ggplot2
scPHO84 <- ggplot(scPHO84_counts, aes(x = condition, y = count)) +
  geom_point(position = position_jitter(width = 0.1, height = 0), size = 3, alpha = 0.8) +
  scale_y_log10(breaks = c(10, 100, 1000, 5000, 100000, 1000000)) +
  labs(title = "PHO84 Expression (YML123C)",
       y = "Normalized count (log scale)",
       x = "Condition") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14)
  )
scPHO84
ggsave(filename = file.path(Scer_plots, "ScPHO84 Expression t0000_vs_t0060.pdf"),
       plot = scPHO84, width = 8, height = 6)
##############################################
# Extract normalized counts for cgPHO84 (gene-GWK60_B02321)
cgPHO84_counts <- plotCounts(cgdds_collapsed,
                           gene = "gene-GWK60_B02321",
                           intgroup = "condition",
                           returnData = TRUE)

# Plot with ggplot2
cgPHO84 <- ggplot(cgPHO84_counts, aes(x = condition, y = count)) +
  geom_point(position = position_jitter(width = 0.1, height = 0), size = 3, alpha = 0.8) +
  scale_y_log10(breaks = c(10, 100, 1000, 5000, 100000, 1000000)) +
  labs(title = "PHO84 Expression (GWK60_B02321)",
       y = "Normalized count (log scale)",
       x = "Condition") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14)
  )
cgPHO84
ggsave(filename = file.path(Cgla_plots, "CgPHO84 Expression t0000_vs_t0060.pdf"),
       plot = cgPHO84, width = 8, height = 6)
################################
# Extract normalized counts for caPHO84
caPHO84_counts <- plotCounts(calbdds_collapsed,
                           gene = "orf19.655",
                           intgroup = "condition",
                           returnData = TRUE)

# Plot with ggplot2
caPHO84 <- ggplot(caPHO84_counts, aes(x = condition, y = count)) +
  geom_point(position = position_jitter(width = 0.1, height = 0), size = 3, alpha = 0.8) +
  scale_y_log10(breaks = c(10, 100, 1000, 5000, 100000, 1000000)) +
  labs(title = "PHO84 Expression (orf19.655)",
       y = "Normalized count (log scale)",
       x = "Condition") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14)
  )
caPHO84
ggsave(filename = file.path(Calb_plots, "CaPHO84 Expression t0000_vs_t0060.pdf"),
       plot = caPHO84, width = 8, height = 6)
########################
# Extract normalized counts for KlPHO84
klPHO84_counts <- plotCounts(klacdds_collapsed,
                           gene = "gene-KDRO_C02880",
                           intgroup = "condition",
                           returnData = TRUE)

# Plot with ggplot2
klPHO84 <- ggplot(klPHO84_counts, aes(x = condition, y = count)) +
  geom_point(position = position_jitter(width = 0.1, height = 0), size = 3, alpha = 0.8) +
  scale_y_log10(breaks = c(10, 100, 1000, 5000, 100000, 1000000)) +
  labs(title = "PHO84 Expression (KDRO_C02880)",
       y = "Normalized count (log scale)",
       x = "Condition") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 14)
  )
klPHO84
ggsave(filename = file.path(Klac_plots, "KlPHO84 Expression t0000_vs_t0060.pdf"),
       plot = klPHO84, width = 8, height = 6)




```


Goal (functional enrichment with topGO)

Identify Biological Process (BP) Gene Ontology terms enriched among DESeq2-tested genes for each species (Scer, Cgla, Calb, Klac), using a consistent pipeline so results are comparable across species and suitable for cross-species summaries.

What this chunk does

Set up: load packages, create per-species plot/output folders, and define small helper functions to save QC plots (p-value histograms, GO term size histograms), barplots comparing topGO algorithms, and GO DAG snapshots.

Gene→GO mappings:
Scer: pull GO annotations from org.Sc.sgd.db (GOALL), merge with GO.db to keep BP only, convert to gene2GO.
Cgla: parse CGD GAF; map CAGL0/GWK60 → QNG via your three-way mapping; keep BP terms; build gene2GO.
Calb: parse CGD GAF; map A21/A22 IDs → orf19; keep BP; build gene2GO.
Klac: parse NCBI GAF; map locus_tag → NCBI GeneID; keep BP; build gene2GO (with a small PHO84/KDRO_C02880 patch so it’s included).

Universe & scores: for each species, take the full DESeq table (*_topGo.csv), set gene-level scores to the DESeq p-value (NAs → 1), and define the selection function p < 0.01.
Run topGO per species: build topGOdata (BP, nodeSize=10), run multiple algorithms (classic, elim, weight01, plus KS for Scer), generate a ranked results table, and write it to disk.
QC & visualization: save p-value and GO size histograms, an algorithm comparison barplot (−log10 p across classic/elim/weight01), and a GO DAG image of the top nodes.
Sanity checks: print annotation coverage (how many genes in the universe are annotated), counts of p=1 terms, distribution of term sizes, and proportion of terms with zero significant genes.
Cross-species summary: combine significant terms (weight01 < 0.05) across species into a presence/absence matrix and save for downstream comparison.

Key parameters & assumptions
Ontology: BP only (can switch to MF/CC with the same framework).
Selection threshold: p < 0.01 for geneSel.
Term size filter: nodeSize = 10.
Input IDs: harmonized per species (Scer ORFs, Cgla QNG, Calb orf19, Klac GeneID/XP) to ensure gene2GO and DESeq universes align.
Outputs: per-species CSV of topGO results, diagnostic PDFs/PNGs, and a cross-species GO comparison CSV.

Why this structure?
Using the complete DESeq universe (not just significant genes) avoids selection bias, topGO’s elim/weight01 reduce GO hierarchy dependencies, and the standardized plotting/QC makes differences across species interpretable and reproducible.

```{r functional-analysis-GO-Term-Enrichment}

############ functional analysis with topGo 
# library(topGO)
# library(biomaRt)
# library(org.Sc.sgd.db)
# library(GO.db)
# library(ggplot2)
# library(tidyr)
# library(stringr)


#plot dirs
# Scer_plots <- here::here("08.plots","Scer"); dir.create(Scer_plots, recursive = TRUE, showWarnings = FALSE)
# Cgla_plots <- here::here("08.plots","Cgla"); dir.create(Cgla_plots, recursive = TRUE, showWarnings = FALSE)
# Calb_plots <- here::here("08.plots","Calb"); dir.create(Calb_plots, recursive = TRUE, showWarnings = FALSE)
# Klac_plots <- here::here("08.plots","Klac"); dir.create(Klac_plots, recursive = TRUE, showWarnings = FALSE)

#helpers to save diagnostics
save_pvalue_hist <- function(result_obj, outfile, title = "P-value distribution") {
  pdf(outfile, width = 7, height = 5)
  pvals <- score(result_obj)
  hist(pvals, breaks = 50, main = title, xlab = "pvals")
  abline(h = length(pvals)/50, col = "red", lty = 2)
  dev.off()
}
save_go_size_hist <- function(GOdata, outfile, title) {
  pdf(outfile, width = 7, height = 5)
  ts <- termStat(GOdata)
  hist(ts$Annotated, breaks = 50, main = title, xlab = "Number of annotated genes")
  dev.off()
}

############ PREPARE UNIFORM PLOTTING HELPERS 
#helpers for uniform plots across species
save_algo_barplot <- function(res_tbl, species, outdir) {
  pd <- res_tbl %>%
    dplyr::slice_head(n = 20) %>%
    dplyr::select(GO.ID, Term, classic, elim, weight01) %>%
    tidyr::pivot_longer(cols = c(classic, elim, weight01),
                        names_to = "Algorithm",
                        values_to = "pvalue") %>%
    dplyr::mutate(pvalue = as.numeric(pvalue),
                  negLog10P = -log10(pvalue + 1e-10),
                  Term = stringr::str_trunc(Term, 40))
  p <- ggplot(pd,
              aes(x = reorder(Term, negLog10P),
                  y = negLog10P,
                  fill = Algorithm)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    coord_flip() +
    labs(title = paste(species, ": topGO Algorithm Comparison"),
         x = NULL, y = "-log10(p-value)") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top",
          axis.text.y = element_text(size = 10)) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed",
               color = "red", alpha = 0.5)
  ggsave(file.path(outdir, paste0(species, "_topGO_algorithm_comparison.pdf")),
         plot = p, width = 10, height = 8, useDingbats = FALSE)
}

save_go_dag <- function(GOdata, result_weight01, species, outdir) {
  #png hi-res
  outfile_png <- file.path(outdir, paste0(species, "_GO_graph_top10.png"))
  grDevices::png(outfile_png,
                 width = 6000, height = 4500,
                 res = 600, type = "cairo", antialias = "subpixel")
  op <- par(mai = c(0, 0, 0, 0))
  try({
    showSigOfNodes(GOdata,
                   score(result_weight01),
                   firstSigNodes = 10,
                   useInfo = "all")
  }, silent = TRUE)
  try(par(op), silent = TRUE)
  try(grDevices::dev.off(), silent = TRUE)

  #pdf via Cairo for crisp vectors
  if (requireNamespace("Cairo", quietly = TRUE)) {
    outfile_pdf <- file.path(outdir, paste0(species, "_GO_graph_top10.pdf"))
    Cairo::CairoPDF(outfile_pdf, width = 16, height = 12)
    op2 <- par(mai = c(0, 0, 0, 0))
    try({
      showSigOfNodes(GOdata,
                     score(result_weight01),
                     firstSigNodes = 10,
                     useInfo = "all")
    }, silent = TRUE)
    try(par(op2), silent = TRUE)
    try(grDevices::dev.off(), silent = TRUE)
  }
}

############PREPARE GENE-TO-GO MAPPINGS

create_gene2GO_mapping <- function(species, gene_ids, annotation_file = NULL) {
  
  if (species == "Scer") {
    #use org.Sc.sgd.db for S. cerevisiae
    #gene_ids already cleaned from scer_topGo
    
    #get GO annotations
    go_map <- select(org.Sc.sgd.db, 
                     keys = gene_ids,
                     columns = c("GOALL", "EVIDENCEALL"),
                     keytype = "ORF")
    
    cat("Initial GO mappings found:", nrow(go_map), "\n")
    
    #remove NA and get GO ontology info
    go_map <- go_map[!is.na(go_map$GOALL), ]
    cat("After removing NA:", nrow(go_map), "\n")
    
    go_bp <- select(GO.db, keys = unique(go_map$GOALL), 
                    columns = "ONTOLOGY", keytype = "GOID")
    go_map <- merge(go_map, go_bp, by.x = "GOALL", by.y = "GOID")
    cat("After merge with GO.db:", nrow(go_map), "\n")
    
    #filter for BP only
    go_map <- go_map[go_map$ONTOLOGY == "BP", ]
    cat("After BP filter:", nrow(go_map), "\n")
    
    #convert to gene2GO list format
    gene2GO <- split(go_map$GOALL, go_map$ORF)
    
  } else if (species == "Cgla") {
    #parse GAF format for C. glabrata from consolidated CGD file
    cat("Parsing GAF file for C. glabrata\n")
    
    #load GWK60 to CAGL0 mapping
    cgla_mapping <- read.csv(here::here("05.metadata", "Cgla", "qng_gwk_cagl_threeway_map_complete.csv"))
    
    #create lookup from CAGL0 to GWK60 for reverse mapping
    cagl_to_gwk <- setNames(cgla_mapping$gwk60_id, cgla_mapping$cagl_id)
    
    #read GAF file
    gaf_data <- read.delim(annotation_file, header = FALSE, comment.char = "!", 
                           stringsAsFactors = FALSE, sep = "\t")
    
    #filter for C. glabrata entries (taxon:284593)
    cgla_entries <- gaf_data[grepl("taxon:284593", gaf_data$V13), ]
    cat("C. glabrata entries found:", nrow(cgla_entries), "\n")
    
    #extract gene2GO mappings
    gene2GO <- list()
    for(i in 1:nrow(cgla_entries)) {
      #extract CAGL0 ID from column 11 (first element before pipe)
      synonyms <- strsplit(cgla_entries$V11[i], "\\|")[[1]]
      cagl_id <- synonyms[1]
      
      #get corresponding GWK60 ID
      gwk_id <- cagl_to_gwk[cagl_id]
      
      #only process if we have a GWK60 mapping and it's in our input list
      if(!is.na(gwk_id) && gwk_id %in% gene_ids) {
        go_term <- cgla_entries$V5[i]
        aspect <- cgla_entries$V9[i]
        
        #only keep BP terms (P = biological process)
        if(aspect == "P") {
          if(gwk_id %in% names(gene2GO)) {
            gene2GO[[gwk_id]] <- c(gene2GO[[gwk_id]], go_term)
          } else {
            gene2GO[[gwk_id]] <- go_term
          }
        }
      }
    }
    
    #remove duplicates
    gene2GO <- lapply(gene2GO, unique)
    cat("Genes with GO annotations:", length(gene2GO), "\n")
    
  } else if (species == "Calb") {
    #parse GAF format for C. albicans from consolidated CGD file
    cat("Parsing GAF file for C. albicans\n")
    
    #load A21/A22 <-> orf19 mapping (columns expected from your builder: a22_id, a21_orf19, a22_orf19)
    calb_mapping <- read.csv(here::here("05.metadata", "Calb", "a21_to_a22_orf19_mapping.csv"),
                             stringsAsFactors = FALSE)
    
    #normalize column names defensively
    names(calb_mapping) <- tolower(names(calb_mapping))
    #we only need: a22_id and an orf19 column; if both a21_orf19 and a22_orf19 exist, they are identical strings
    if (!("a22_id" %in% names(calb_mapping)) ||
        !any(c("a21_orf19","a22_orf19","orf19") %in% names(calb_mapping))) {
      stop("Calb mapping file must contain a22_id and an orf19 column (a21_orf19 / a22_orf19 / orf19).")
    }
    if (!("orf19" %in% names(calb_mapping))) {
      calb_mapping$orf19 <- if ("a21_orf19" %in% names(calb_mapping)) calb_mapping$a21_orf19 else calb_mapping$a22_orf19
    }
    calb_mapping <- unique(calb_mapping[, c("a22_id","orf19")])
    
    #build lookups to map your input IDs (which may be orf19.* or A22 IDs) to orf19 IDs
    orf19_to_orf19 <- setNames(calb_mapping$orf19, calb_mapping$orf19)   # identity
    a22_to_orf19   <- setNames(calb_mapping$orf19, calb_mapping$a22_id)  # A22 -> orf19
    
    #convert input gene IDs to orf19 IDs (keep original ids for naming gene2GO later)
    input <- gene_ids
    is_orf19 <- grepl("^orf19\\.[0-9]+$", input)
    mapped_orf19 <- rep(NA_character_, length(input))
    mapped_orf19[is_orf19]  <- orf19_to_orf19[input[is_orf19]]
    mapped_orf19[!is_orf19] <- a22_to_orf19[input[!is_orf19]]
    
    keep <- !is.na(mapped_orf19)
    input_kept <- input[keep]
    orf19_ids  <- mapped_orf19[keep]
    
    cat("Input genes:", length(input), "\n")
    cat("Successfully mapped to orf19:", length(orf19_ids), "\n")
    
    #read GAF file
    gaf_data <- read.delim(annotation_file, header = FALSE, comment.char = "!", 
                           stringsAsFactors = FALSE, sep = "\t")
    
    #filter for C. albicans entries (taxon:237561)
    calb_entries <- gaf_data[grepl("taxon:237561", gaf_data$V13), ]
    cat("C. albicans entries found:", nrow(calb_entries), "\n")
    
    #tokenize synonyms; include V2 (DB_Object_ID) itself as a token for matching
    split_syns <- function(s) unlist(strsplit(s, "\\|"), use.names = FALSE)
    tokens_by_row <- lapply(seq_len(nrow(calb_entries)), function(i) {
      syns <- calb_entries$V11[i]
      dbid <- calb_entries$V2[i]
      if (is.na(syns) || syns == "") return(dbid)
      c(dbid, split_syns(syns))
    })
    
    #extract gene2GO mappings keyed by your ORIGINAL input IDs
    gene2GO <- vector("list", length = 0)
    for (i in seq_len(nrow(calb_entries))) {
      aspect <- calb_entries$V9[i]
      if (aspect != "P") next  #keep BP only
      
      go_term <- calb_entries$V5[i]
      toks <- tokens_by_row[[i]]
      
      #does this row carry an orf19 we mapped to?
      hit_orf19 <- toks[grepl("^orf19\\.[0-9]+$", toks)]
      if (length(hit_orf19) == 0) next
      
      #which of our inputs map to this orf19?
      matched_inputs <- input_kept[orf19_ids %in% hit_orf19]
      if (length(matched_inputs) == 0) next
      
      for (gid in matched_inputs) {
        if (is.null(gene2GO[[gid]])) {
          gene2GO[[gid]] <- go_term
        } else {
          gene2GO[[gid]] <- c(gene2GO[[gid]], go_term)
        }
      }
    }
    
    #deduplicate & ensure character vectors
    if (length(gene2GO) > 0) {
      gene2GO <- lapply(gene2GO, function(x) unique(as.character(x)))
      #drop any empties just in case
      gene2GO <- gene2GO[lengths(gene2GO) > 0]
    }
    
    cat("Genes with GO annotations (Calb):", length(gene2GO), "\n")
    
    
  } else if (species == "Klac") {
    #parse GAF format for K. lactis
    cat("Parsing GAF file for K. lactis\n")
    
    #read your locus_tag <-> geneid map
    lt_map <- read.csv(here::here("05.metadata","Klac","klac_locus_tag_to_geneid.csv"),
                       stringsAsFactors = FALSE)
    names(lt_map) <- tolower(names(lt_map))
    stopifnot(all(c("locus_tag","geneid") %in% names(lt_map)))
    lt2gid <- setNames(lt_map$geneid, lt_map$locus_tag)
    
    #map input ids (klla0_*) to ncbi geneid
    input <- gene_ids
    is_num <- grepl("^[0-9]+$", input)
    input_gid <- rep(NA_character_, length(input))
    input_gid[is_num]  <- input[is_num]
    input_gid[!is_num] <- lt2gid[input[!is_num]]
    
    keep <- !is.na(input_gid)
    if (!any(keep)) { warning("K. lactis: no IDs mapped to GeneID via klac_locus_tag_to_geneid.csv"); return(list()) }
    input_gid <- input_gid[keep]
    input_orig <- input[keep]
    
    #read GAF and keep k. lactis bp rows
    gaf_data <- read.delim(annotation_file, header = FALSE, comment.char = "!", 
                           stringsAsFactors = FALSE, sep = "\t")
    #columns: 2=db_object_id (geneid), 5=go id, 9=aspect, 13=taxon
    colnames(gaf_data)[c(2,5,9,13)] <- c("db_obj_id","go_id","aspect","taxon")
    gaf_data <- gaf_data[grepl("taxon:28985", gaf_data$taxon) & gaf_data$aspect == "P", c("db_obj_id","go_id")]
    if (nrow(gaf_data) == 0) { warning("K. lactis: GAF has no BP rows for taxon:28985"); return(list()) }
    gaf_data$db_obj_id <- as.character(gaf_data$db_obj_id)
    gaf_data$go_id <- trimws(gaf_data$go_id)
    
    #subset to mapped geneids and split
    sub <- gaf_data[gaf_data$db_obj_id %in% input_gid, , drop = FALSE]
    if (nrow(sub) == 0) { warning("K. lactis: no overlap between mapped GeneIDs and GAF"); return(list()) }
    gid2go <- split(sub$go_id, sub$db_obj_id)
    
    #rekey to your original ids (klla0_*)
    gid2orig <- split(input_orig, input_gid)
    gene2GO <- vector("list", 0)
    for (gid in names(gid2go)) {
      gos <- unique(as.character(gid2go[[gid]]))
      if (!is.null(gid2orig[[gid]])) {
        for (orig in gid2orig[[gid]]) gene2GO[[orig]] <- gos
      }
    }
    gene2GO <- gene2GO[lengths(gene2GO) > 0]
    cat("Genes with GO annotations:", length(gene2GO), "\n")
  }
  return(gene2GO)
}



############RUN topGO FOR S. CEREVISIAE 

#prepare gene universe and scores
scer_universe <- scer_topGo$gene_id
#check what types of gene IDs we have
cat("Total genes:", length(scer_universe), "\n")
cat("Sample gene IDs:", head(scer_universe, 20), "\n")
cat("Y-genes:", sum(grepl("^Y", scer_universe)), "\n")
cat("Q-genes:", sum(grepl("^Q", scer_universe)), "\n")

#only use Y-genes that org.Sc.sgd.db recognizes
scer_y_genes <- scer_topGo[grepl("^Y", scer_topGo$gene_id), ]
scer_pvalues <- setNames(scer_y_genes$pvalue, scer_y_genes$gene_id)
scer_pvalues[is.na(scer_pvalues)] <- 1
#define selection function
topDiffGenes <- function(allScore) {
  return(allScore < 0.01)
}

#get gene2GO mapping for S. cerevisiae
scer_gene2GO <- create_gene2GO_mapping("Scer", names(scer_pvalues))

#debug: check if gene2GO mapping is populated
cat("Number of genes in gene2GO:", length(scer_gene2GO), "\n")
cat("First few gene IDs in gene2GO:", head(names(scer_gene2GO)), "\n")
cat("First few gene IDs in scer_pvalues:", head(names(scer_pvalues)), "\n")
cat("Sample GO terms for first gene:", head(scer_gene2GO[[1]]), "\n")

#check for overlap
cat("Genes in both:", length(intersect(names(scer_pvalues), names(scer_gene2GO))), "\n")

#build topGOdata object
GOdata_Scer <- new("topGOdata",
                   description = "S. cerevisiae Pi starvation 60min",
                   ontology = "BP",
                   allGenes = scer_pvalues,
                   geneSel = topDiffGenes,
                   annot = annFUN.gene2GO,
                   gene2GO = scer_gene2GO,
                   nodeSize = 10)

#run multiple algorithms
#classic Fisher
result_Scer_classicFisher <- runTest(GOdata_Scer, 
                                     algorithm = "classic", 
                                     statistic = "fisher")

#classic KS
result_Scer_classicKS <- runTest(GOdata_Scer, 
                                 algorithm = "classic", 
                                 statistic = "ks")

#elim Fisher
result_Scer_elimFisher <- runTest(GOdata_Scer, 
                                  algorithm = "elim", 
                                  statistic = "fisher")

#weight01 Fisher (topGO recommended)
result_Scer_weight01 <- runTest(GOdata_Scer, 
                                algorithm = "weight01", 
                                statistic = "fisher")

#generate results table
Scer_topGO_results <- GenTable(GOdata_Scer,
                               classic = result_Scer_classicFisher,
                               elim = result_Scer_elimFisher,
                               weight01 = result_Scer_weight01,
                               KS = result_Scer_classicKS,
                               orderBy = "weight01",
                               ranksOf = "classic",
                               topNodes = 30)

#save results
write.csv(Scer_topGO_results, 
         file.path(Scer_results, "Scer_topGO_complete.csv"),
         row.names = FALSE)

############VISUALIZATION 

library(ggplot2)
library(tidyr)

#prepare data for plotting
plot_data <- Scer_topGO_results %>%
  slice_head(n = 20) %>%
  dplyr::select(GO.ID, Term, classic, elim, weight01) %>%
  pivot_longer(cols = c(classic, elim, weight01),
               names_to = "Algorithm",
               values_to = "pvalue") %>%
  mutate(
    pvalue = as.numeric(pvalue),
    negLog10P = -log10(pvalue + 1e-10),
    Term = str_trunc(Term, 40)
  )

#create comparison plot
p_topGO_compare <- ggplot(plot_data, 
                          aes(x = reorder(Term, negLog10P), 
                              y = negLog10P, 
                              fill = Algorithm)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  coord_flip() +
  labs(title = "S. cerevisiae: topGO Algorithm Comparison",
       x = NULL,
       y = "-log10(p-value)") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    axis.text.y = element_text(size = 10)
  ) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = "dashed", 
             color = "red", 
             alpha = 0.5)

print(p_topGO_compare)

#save bar plot
ggsave(file.path(Scer_plots, "Scer_topGO_algorithm_comparison.pdf"),
       plot = p_topGO_compare,
       width = 10, height = 8)

#png of DAG 
if (!ok || !file.exists(outfile_svg) || file.info(outfile_svg)$size <= 1000) {
  # 1) high-res PNG for slides/docs (crisp on zoom)
  outfile_png <- file.path(Scer_plots, "Scer_GO_graph_top10.png")
  grDevices::png(outfile_png,
                 width = 6000, height = 4500,    # big canvas
                 res = 600,                      # high DPI
                 type = "cairo",                 # better text/lines on Windows
                 antialias = "subpixel")
  op <- par(mai = c(0, 0, 0, 0))                 # use full page
  on.exit({ try(par(op), silent = TRUE); try(grDevices::dev.off(), silent = TRUE) }, add = TRUE)
  showSigOfNodes(
    GOdata_Scer,
    score(result_Scer_weight01),
    firstSigNodes = 10,
    useInfo = "all"
  )

  #PDF via Cairo here for better zoom
  if (requireNamespace("Cairo", quietly = TRUE)) {
    outfile_pdf <- file.path(Scer_plots, "Scer_GO_graph_top10.pdf")
    Cairo::CairoPDF(outfile_pdf, width = 16, height = 12)  # larger page for legibility
    op2 <- par(mai = c(0, 0, 0, 0))
    try({
      showSigOfNodes(
        GOdata_Scer,
        score(result_Scer_weight01),
        firstSigNodes = 10,
        useInfo = "all"
      )
    }, silent = TRUE)
    try(par(op2), silent = TRUE)
    try(grDevices::dev.off(), silent = TRUE)
  }
}

# (ADDED) UNIFORM PLOTS FOR SCER
save_algo_barplot(Scer_topGO_results, "Scer", Scer_plots)
save_go_dag(GOdata_Scer, result_Scer_weight01, "Scer", Scer_plots)

############SANITY CHECKS

#check annotation coverage
check_annotation_coverage <- function(GOdata) {
  #get total genes from slot
  total_genes <- length(GOdata@allGenes)
  annotated <- numGenes(GOdata)
  sig_genes_total <- numSigGenes(GOdata)
  
  #get significant genes that are annotated
  sig_genes_list <- sigGenes(GOdata)
  sig_annotated <- length(sig_genes_list)
  
  cat("Annotation Coverage:\n")
  cat("  Total genes in universe:", total_genes, "\n")
  cat("  Annotated genes:", annotated, 
      sprintf("(%.1f%%)\n", 100 * annotated/total_genes))
  cat("  Significant genes (in universe):", sig_genes_total, "\n")
  cat("  Significant genes (annotated):", sig_annotated, "\n")
  cat("  GO terms tested:", length(usedGO(GOdata)), "\n")
}
check_annotation_coverage(GOdata_Scer)

#check p-value distributions
check_pvalue_dist <- function(result_obj) {
  pvals <- score(result_obj)
  hist(pvals, breaks = 50, main = "P-value distribution")
  abline(h = length(pvals)/50, col = "red", lty = 2)
}

check_pvalue_dist(result_Scer_classicFisher)

#save diagnostics for scer
save_pvalue_hist(result_Scer_classicFisher,
                 file.path(Scer_plots, "Scer_pvalue_distribution.pdf"),
                 "P-value distribution - S. cerevisiae")
save_go_size_hist(GOdata_Scer,
                  file.path(Scer_plots, "Scer_GO_term_sizes.pdf"),
                  "GO Term Size Distribution - S. cerevisiae")

#investigate p=1 GO terms
p1_terms <- Scer_topGO_results[Scer_topGO_results$classic == "1", ]
cat("Number of GO terms with p=1:", nrow(p1_terms), "\n")

#check their characteristics
if(nrow(p1_terms) > 0) {
  cat("Sample terms with p=1:\n")
  print(p1_terms[1:min(5, nrow(p1_terms)), c("GO.ID", "Term", "Annotated", "Significant", "Expected")])
}

#get distribution of GO term sizes
all_go_sizes <- termStat(GOdata_Scer)
hist(all_go_sizes$Annotated, breaks=50, main="GO Term Size Distribution", xlab="Number of annotated genes")

#check terms with 0 significant genes
zero_sig <- all_go_sizes[all_go_sizes$Significant == 0, ]
cat("GO terms with 0 significant genes:", nrow(zero_sig), "\n")
cat("Proportion of all terms:", round(100*nrow(zero_sig)/nrow(all_go_sizes), 1), "%\n")

############OTHER SPECIES ANALYSIS

#define annotation file paths - using consolidated GAF for Candida species
cgd_gaf <- here::here("02.annotationdata", "go-terms", "gene_association.cgd")
klac_gaf <- here::here("02.annotationdata", "go-terms", "GCF_000002515.2_ASM251v1_gene_ontology.gaf")

############C. GLABRATA
cat("\n========== C. glabrata topGO Analysis ==========\n")

#prepare gene universe and scores
cgla_universe <- cgla_topGo$gene_id
cgla_pvalues <- setNames(cgla_topGo$pvalue, cgla_topGo$gene_id)
cgla_pvalues[is.na(cgla_pvalues)] <- 1

#get gene2GO mapping
cgla_gene2GO <- create_gene2GO_mapping("Cgla", names(cgla_pvalues), cgd_gaf)

#build topGOdata object
GOdata_Cgla <- new("topGOdata",
                   description = "C. glabrata Pi starvation 60min",
                   ontology = "BP",
                   allGenes = cgla_pvalues,
                   geneSel = topDiffGenes,
                   annot = annFUN.gene2GO,
                   gene2GO = cgla_gene2GO,
                   nodeSize = 10)

#run algorithms
result_Cgla_classicFisher <- runTest(GOdata_Cgla, algorithm = "classic", statistic = "fisher")
result_Cgla_elimFisher <- runTest(GOdata_Cgla, algorithm = "elim", statistic = "fisher")
result_Cgla_weight01 <- runTest(GOdata_Cgla, algorithm = "weight01", statistic = "fisher")

#generate results table
Cgla_topGO_results <- GenTable(GOdata_Cgla,
                               classic = result_Cgla_classicFisher,
                               elim = result_Cgla_elimFisher,
                               weight01 = result_Cgla_weight01,
                               orderBy = "weight01",
                               ranksOf = "classic",
                               topNodes = 30)

#save results
write.csv(Cgla_topGO_results, 
         file.path(Cgla_results, "Cgla_topGO_complete.csv"),
         row.names = FALSE)

#check coverage
check_annotation_coverage(GOdata_Cgla)

#check p-value distributions
check_pvalue_dist(result_Cgla_classicFisher)

#save diagnostics for cgla
save_pvalue_hist(result_Cgla_classicFisher,
                 file.path(Cgla_plots, "Cgla_pvalue_distribution.pdf"),
                 "P-value distribution - C. glabrata")
save_go_size_hist(GOdata_Cgla,
                  file.path(Cgla_plots, "Cgla_GO_term_sizes.pdf"),
                  "GO Term Size Distribution - C. glabrata")

# (ADDED) UNIFORM PLOTS FOR CGLA
save_algo_barplot(Cgla_topGO_results, "Cgla", Cgla_plots)
save_go_dag(GOdata_Cgla, result_Cgla_weight01, "Cgla", Cgla_plots)

#investigate p=1 GO terms
p1_terms_cgla <- Cgla_topGO_results[Cgla_topGO_results$classic == "1", ]
cat("Number of GO terms with p=1:", nrow(p1_terms_cgla), "\n")

#check their characteristics
if(nrow(p1_terms_cgla) > 0) {
  cat("Sample terms with p=1:\n")
  print(p1_terms_cgla[1:min(5, nrow(p1_terms_cgla)), c("GO.ID", "Term", "Annotated", "Significant", "Expected")])
}

#get distribution of GO term sizes
all_go_sizes_cgla <- termStat(GOdata_Cgla)
hist(all_go_sizes_cgla$Annotated, breaks=50, main="GO Term Size Distribution - C. glabrata", xlab="Number of annotated genes")

#check terms with 0 significant genes
zero_sig_cgla <- all_go_sizes_cgla[all_go_sizes_cgla$Significant == 0, ]
cat("GO terms with 0 significant genes:", nrow(zero_sig_cgla), "\n")
cat("Proportion of all terms:", round(100*nrow(zero_sig_cgla)/nrow(all_go_sizes_cgla), 1), "%\n")

############C. ALBICANS
cat("\n========== C. albicans topGO Analysis ==========\n")

#prepare gene universe and scores
calb_universe <- calb_topGo$gene_id
calb_pvalues <- setNames(calb_topGo$pvalue, calb_topGo$gene_id)
calb_pvalues[is.na(calb_pvalues)] <- 1

#get gene2GO mapping
calb_gene2GO <- create_gene2GO_mapping("Calb", names(calb_pvalues), cgd_gaf)

#build topGOdata object
GOdata_Calb <- new("topGOdata",
                   description = "C. albicans Pi starvation 60min",
                   ontology = "BP",
                   allGenes = calb_pvalues,
                   geneSel = topDiffGenes,
                   annot = annFUN.gene2GO,
                   gene2GO = calb_gene2GO,
                   nodeSize = 10)

#run algorithms
result_Calb_classicFisher <- runTest(GOdata_Calb, algorithm = "classic", statistic = "fisher")
result_Calb_elimFisher <- runTest(GOdata_Calb, algorithm = "elim", statistic = "fisher")
result_Calb_weight01 <- runTest(GOdata_Calb, algorithm = "weight01", statistic = "fisher")

#generate results table
Calb_topGO_results <- GenTable(GOdata_Calb,
                               classic = result_Calb_classicFisher,
                               elim = result_Calb_elimFisher,
                               weight01 = result_Calb_weight01,
                               orderBy = "weight01",
                               ranksOf = "classic",
                               topNodes = 30)

#save results
write.csv(Calb_topGO_results, 
         file.path(Calb_results, "Calb_topGO_complete.csv"),
         row.names = FALSE)

#check coverage
check_annotation_coverage(GOdata_Calb)

#check p-value distributions
check_pvalue_dist(result_Calb_classicFisher)

#save diagnostics for calb
save_pvalue_hist(result_Calb_classicFisher,
                 file.path(Calb_plots, "Calb_pvalue_distribution.pdf"),
                 "P-value distribution - C. albicans")
save_go_size_hist(GOdata_Calb,
                  file.path(Calb_plots, "Calb_GO_term_sizes.pdf"),
                  "GO Term Size Distribution - C. albicans")

# (ADDED) UNIFORM PLOTS FOR CALB
save_algo_barplot(Calb_topGO_results, "Calb", Calb_plots)
save_go_dag(GOdata_Calb, result_Calb_weight01, "Calb", Calb_plots)

#investigate p=1 GO terms
p1_terms_calb <- Calb_topGO_results[Calb_topGO_results$classic == "1", ]
cat("Number of GO terms with p=1:", nrow(p1_terms_calb), "\n")

#check their characteristics
if(nrow(p1_terms_calb) > 0) {
  cat("Sample terms with p=1:\n")
  print(p1_terms_calb[1:min(5, nrow(p1_terms_calb)), c("GO.ID", "Term", "Annotated", "Significant", "Expected")])
}

#get distribution of GO term sizes
all_go_sizes_calb <- termStat(GOdata_Calb)
hist(all_go_sizes_calb$Annotated, breaks=50, main="GO Term Size Distribution - C. albicans", xlab="Number of annotated genes")

#check terms with 0 significant genes
zero_sig_calb <- all_go_sizes_calb[all_go_sizes_calb$Significant == 0, ]
cat("GO terms with 0 significant genes:", nrow(zero_sig_calb), "\n")
cat("Proportion of all terms:", round(100*nrow(zero_sig_calb)/nrow(all_go_sizes_calb), 1), "%\n")

############K. LACTIS
cat("\n========== K. lactis topGO Analysis ==========\n")

#prepare gene universe and scores
klac_universe <- klac_topGo$gene_id
klac_pvalues <- setNames(klac_topGo$pvalue, klac_topGo$gene_id)
klac_pvalues[is.na(klac_pvalues)] <- 1

#get gene2GO mapping
klac_gene2GO <- create_gene2GO_mapping("Klac", names(klac_pvalues), klac_gaf)

#KDRO_C02880 is in universe (annotation patch)
if (!"KDRO_C02880" %in% names(klac_pvalues)) {
  klac_pvalues <- c(klac_pvalues, setNames(1, "KDRO_C02880"))
}
#bp term for pho84/kdro_c02880
klac_gene2GO[["KDRO_C02880"]] <- unique(c(klac_gene2GO[["KDRO_C02880"]], "GO:0006817"))

#overlap guard so topGO sees real terms
klac_overlap <- intersect(names(klac_pvalues), names(klac_gene2GO))
cat("K. lactis overlap (universe ∩ gene2GO):", length(klac_overlap), "\n")
klac_pvalues_use <- klac_pvalues[klac_overlap]
klac_gene2GO_use <- klac_gene2GO[klac_overlap]
stopifnot(length(klac_pvalues_use) > 0, length(klac_gene2GO_use) > 0)

#build topgo object
GOdata_Klac <- new("topGOdata",
                   description = "K. lactis Pi starvation 60min",
                   ontology = "BP",
                   allGenes = klac_pvalues_use,
                   geneSel = topDiffGenes,
                   annot = annFUN.gene2GO,
                   gene2GO = klac_gene2GO_use,
                   nodeSize = 10)

#run algorithms
result_Klac_classicFisher <- runTest(GOdata_Klac, algorithm = "classic", statistic = "fisher")
result_Klac_elimFisher <- runTest(GOdata_Klac, algorithm = "elim", statistic = "fisher")
result_Klac_weight01 <- runTest(GOdata_Klac, algorithm = "weight01", statistic = "fisher")

#generate results table
Klac_topGO_results <- GenTable(GOdata_Klac,
                               classic = result_Klac_classicFisher,
                               elim = result_Klac_elimFisher,
                               weight01 = result_Klac_weight01,
                               orderBy = "weight01",
                               ranksOf = "classic",
                               topNodes = 30)

#save results
write.csv(Klac_topGO_results, 
         file.path(Klac_results, "Klac_topGO_complete.csv"),
         row.names = FALSE)

#check coverage
check_annotation_coverage(GOdata_Klac)

#check p-value distributions
check_pvalue_dist(result_Klac_classicFisher)

#save diagnostics for klac
save_pvalue_hist(result_Klac_classicFisher,
                 file.path(Klac_plots, "Klac_pvalue_distribution.pdf"),
                 "P-value distribution - K. lactis")
save_go_size_hist(GOdata_Klac,
                  file.path(Klac_plots, "Klac_GO_term_sizes.pdf"),
                  "GO Term Size Distribution - K. lactis")

# (ADDED) UNIFORM PLOTS FOR KLAC
save_algo_barplot(Klac_topGO_results, "Klac", Klac_plots)
save_go_dag(GOdata_Klac, result_Klac_weight01, "Klac", Klac_plots)

#investigate p=1 GO terms
p1_terms_klac <- Klac_topGO_results[Klac_topGO_results$classic == "1", ]
cat("Number of GO terms with p=1:", nrow(p1_terms_klac), "\n")

#check their characteristics
if(nrow(p1_terms_klac) > 0) {
  cat("Sample terms with p=1:\n")
  print(p1_terms_klac[1:min(5, nrow(p1_terms_klac)), c("GO.ID", "Term", "Annotated", "Significant", "Expected")])
}

#get distribution of GO term sizes
all_go_sizes_klac <- termStat(GOdata_Klac)
hist(all_go_sizes_klac$Annotated, breaks=50, main="GO Term Size Distribution - K. lactis", xlab="Number of annotated genes")

#check terms with 0 significant genes
zero_sig_klac <- all_go_sizes_klac[all_go_sizes_klac$Significant == 0, ]
cat("GO terms with 0 significant genes:", nrow(zero_sig_klac), "\n")
cat("Proportion of all terms:", round(100*nrow(zero_sig_klac)/nrow(all_go_sizes_klac), 1), "%\n")

############CROSS-SPECIES COMPARISON
cat("\n========== Cross-species topGO Comparison ==========\n")

#compile all results
all_species_results <- list(
  Scer = Scer_topGO_results,
  Cgla = Cgla_topGO_results,
  Calb = Calb_topGO_results,
  Klac = Klac_topGO_results
)

#extract significant terms from each species
significant_terms <- lapply(names(all_species_results), function(sp) {
  df <- all_species_results[[sp]]
  df$Species <- sp
  df[as.numeric(df$weight01) < 0.05, c("GO.ID", "Term", "Species", "weight01")]
})

#combine and create comparison matrix
combined_terms <- dplyr::bind_rows(significant_terms)
term_matrix <- combined_terms %>%
  dplyr::select(GO.ID, Term, Species) %>%
  dplyr::mutate(present = 1) %>%
  tidyr::pivot_wider(names_from = Species, 
              values_from = present, 
              values_fill = 0)

#save comparison
write.csv(term_matrix, 
         file.path(here::here("06.results"), "Cross_species_GO_comparison.csv"),
         row.names = FALSE)

#print summary
cat("\nShared GO terms across species:\n")
shared_all <- term_matrix[rowSums(term_matrix[,3:6]) == 4, ]
cat("Shared by all 4 species:", nrow(shared_all), "\n")
if(nrow(shared_all) > 0) {
  print(head(shared_all$Term))
}
```


```{r orthogroup-Assignment, warning=FALSE, message=FALSE}
############################# S.cerevisiae, classify the whole genome into either single core, multicore, none-core genes
# Three classes per species (based on copy number + presence across the compared species):
# Single-core: a single-copy gene in the focal species with orthologs in the other species being compared. 

# Multi-core: a duplicated (multi-copy) gene in the focal species with orthologs in the other species. 

# Non-core: no ortholog detected in at least one of the other species (not shared with ). 

# ---- extracting ID patterns with regexp----

scer_pat <- "\\bY[A-P][LR][0-9]{3}[WC](?:-[A-Z])?\\b"   # SGD ORFs
cgla_pat <- "QNG\\d+(?:\\.\\d+)?"                      # Cg: QNG only
calb_pat <- "orf\\d+\\.\\d+"                           # Ca: orf19.xxxx
klac_pat <- "XP_\\d+(?:\\.\\d+)?"                      # Kl: RefSeq XP_

collapse_or_na <- function(v) { v <- unique(v); if (length(v)) paste(v, collapse=";") else NA_character_ }

# Core-status helper
core_status <- function(n_self, n_o1, n_o2, n_o3){
  dplyr::case_when(
    n_self == 0                        ~ NA_character_,   # focal absent
    n_o1 == 0 | n_o2 == 0 | n_o3 == 0 ~ "non-core",      # not shared with all others
    n_self >= 2                        ~ "multi-core",    # shared, self has 2+
    n_self == 1                        ~ "single-core"    # shared, self has 1
  )
}

# ================== SLIMMED ORTHOGROUP TABLE ==================
orthogroups_cleaned <- orthogroups %>%
  select(OG, S_cerevisiae, C_glabrata, C_albicans, K_lactis) %>%
  mutate(across(c(S_cerevisiae, C_glabrata, C_albicans, K_lactis),
                ~ na_if(trimws(.), ""))) %>%

  # Extract all gene IDs per species (list-columns)
  mutate(
    scer_ids = str_extract_all(replace_na(S_cerevisiae, ""), scer_pat) %>% lapply(unique),
    cgla_ids = str_extract_all(replace_na(C_glabrata,  ""), cgla_pat) %>% lapply(unique),
    calb_ids = str_extract_all(replace_na(C_albicans,  ""), calb_pat) %>% lapply(unique),
    klac_ids = str_extract_all(replace_na(K_lactis,    ""), klac_pat) %>% lapply(unique),

    # Flatten to readable lists for each orthogroup
    Scer_orthgrp_list = vapply(scer_ids, collapse_or_na, character(1)),
    Cgla_orthgrp_list = vapply(cgla_ids, collapse_or_na, character(1)),
    Calb_orthgrp_list = vapply(calb_ids, collapse_or_na, character(1)),
    Klac_orthgrp_list = vapply(klac_ids, collapse_or_na, character(1)),

    # Copy counts per species (rename to n_orthgrp_* as requested)
    n_orthgrp_Scer = lengths(scer_ids),
    n_orthgrp_Cgla = lengths(cgla_ids),
    n_orthgrp_Calb = lengths(calb_ids),
    n_orthgrp_Klac = lengths(klac_ids)
  ) %>%

  # Core status for EACH focal species using the n_orthgrp_* counts
  mutate(
    scer_core_status = core_status(n_orthgrp_Scer, n_orthgrp_Cgla, n_orthgrp_Calb, n_orthgrp_Klac),
    cgla_core_status = core_status(n_orthgrp_Cgla, n_orthgrp_Scer, n_orthgrp_Calb, n_orthgrp_Klac),
    calb_core_status = core_status(n_orthgrp_Calb, n_orthgrp_Scer, n_orthgrp_Cgla, n_orthgrp_Klac),
    klac_core_status = core_status(n_orthgrp_Klac, n_orthgrp_Scer, n_orthgrp_Cgla, n_orthgrp_Calb)
  ) %>%

  # Keep only the relevanr columns 
  select(
    OG,
    S_cerevisiae, Scer_orthgrp_list, n_orthgrp_Scer, scer_core_status,
    C_glabrata,   Cgla_orthgrp_list, n_orthgrp_Cgla, cgla_core_status,
    C_albicans,   Calb_orthgrp_list, n_orthgrp_Calb, calb_core_status,
    K_lactis,     Klac_orthgrp_list, n_orthgrp_Klac, klac_core_status
  )

####### write into csv file 
write_csv(orthogroups_cleaned, file.path(combined, "20250910_orthogroups_cleaned.csv"))

####################### visualize results 
# Long table of per-species core status
status_long <- orthogroups_cleaned %>%
  select(scer_core_status, cgla_core_status, calb_core_status, klac_core_status) %>%
  tidyr::pivot_longer(everything(), names_to = "species", values_to = "status") %>%
  mutate(
    species = recode(species,
      scer_core_status = "Scer",
      cgla_core_status = "Cgla",
      calb_core_status = "Calb",
      klac_core_status = "Klac"
    ),
    status  = fct_explicit_na(status, na_level = "unmapped"),
    status  = factor(status, levels = c("non-core","multi-core","single-core","unmapped")),
    species = factor(species, levels = c("Scer","Cgla","Calb","Klac"))
  )

# Counts per status × species (includes "unmapped")
counts <- status_long %>% 
  count(status, species, name = "n")

# # (optional) percent per species
counts_pct <- counts %>%
  group_by(species) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

# Plot: 4 bars per status (grouped by species)
############ set theme 
bold_theme <- theme(
  axis.title.x = element_text(face = "bold"),
  axis.title.y = element_text(face = "bold"),
  axis.text.x  = element_text(face = "bold"),
  axis.text.y  = element_text(face = "bold"),
  plot.title   = element_text(face = "bold"),
  legend.title = element_text(face = "bold"),
  legend.text  = element_text(face = "bold")
) 
  
p_counts <- ggplot(counts, aes(x = status, y = n, fill = species)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = n),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3, fontface = "bold") +
  labs(x = NULL, y = "Number of genes", title = "Core status by species") +
  theme_minimal() + bold_theme

# # (optional) Percent version
p_pct <- ggplot(counts_pct, aes(x = status, y = pct, fill = species)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", pct)),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3, fontface = "bold") +
  labs(x = NULL, y = "Percent of genes per species", title = "Core status (percent) by species") +
  theme_minimal() + bold_theme

 p_counts
 p_pct  # uncomment to see percentage plot

##### If you also want thicker axis lines/ticks (not just bold text), add:
# bold_theme <- bold_theme + theme(
#   axis.line  = element_line(linewidth = 0.7),
#   axis.ticks = element_line(linewidth = 0.7)
# )

# Load Scer DEGs results 
Scer_Degs  <- read_csv(here("06.results", "Scer", "20250826Scer.LFCgt2.padj0.01.csv"), 
                          show_col_types = FALSE) %>%
  rename(scer_gene_id = gene_id)

# Left join to preserve all DEGs
Scer_og <- Scer_Degs %>% 
  left_join(orthogroup2, by = "scer_gene_id") #%>%
  #select(Scer_geneID, OG, C_albicans, C_glabrata, K_lactis, S_cerevisiae)


############################ C.glabrata

# Load Cgla DEGs results 
Cgla_Degs  <- read_csv(here("06.results", "Cgla", "20250826Cgla.LFCgt2.padj0.01.csv"), 
                       show_col_types = FALSE) #%>%
  #rename(gwk60_id = gene_id)

############ read in the C.glabrata proteom to BG2 annotation mapping file 
Cgla_BG2Protmapping  <- read_csv(here("06.results", "Cgla", "qng_gwk_cagl_threeway_map_complete.csv"), 
                                 show_col_types = FALSE) %>%
  rename(gene_id = gwk60_id)

###################  Map IDs in DEGs file to the proteom IDs used in orthofinder 
Cgla_mapped <- Cgla_Degs %>%
  #distinct(gwk60_id) %>%
  left_join(Cgla_BG2Protmapping,
            by = "gene_id") %>%
  rename(cgla_gene_id = qng_id)

# Left join to preserve all DEGs
Cgla_og <- Cgla_mapped %>%
  left_join(
    orthogroup2 %>% 
      filter(!is.na(cgla_gene_id)) %>% 
      distinct(cgla_gene_id, .keep_all = TRUE),
    by = "cgla_gene_id"
  ) #%>%
  #select(qng_id, gwk60_id, cagl_id,gene_name, OG, C_albicans, C_glabrata, K_lactis, S_cerevisiae)

############### C. albicans 
# Load Cgla DEGs results 
Calb_Degs  <- read_csv(here("06.results", "Calb", "20250826Calb.LFCgt2.padj0.01.csv"), 
                       show_col_types = FALSE) %>% 
  rename(calb_gene_id = gene_id)

######### left join
Calb_og <- Calb_Degs %>%
  left_join(
    orthogroup2 %>% 
      filter(!is.na(calb_gene_id)) %>% 
      distinct(calb_gene_id, .keep_all = TRUE),
    by = "calb_gene_id"
  ) #%>%


################## K.lactis  
# Load Klac DEGs results 
Klac_Degs  <- read_csv(here("06.results", "Klac", "20250826Klac.LFCgt2.padj0.01.csv"), 
                       show_col_types = FALSE) %>%
  rename(locus_tag = gene_id)

############ read in the C.glabrata proteom to BG2 annotation mapping file 
Klac_xpklla0tmapping  <- read_csv(here("06.results", "Klac", "xp_to_klla0_mapping.csv"), 
                                  show_col_types = FALSE) 
  

#### left join to Degs 
Klac_mapped <- Klac_Degs %>%
  #distinct(gwk60_id) %>%
  left_join(Klac_xpklla0tmapping,
            by = "locus_tag") %>%
  rename(klac_gene_id = xp_id)

# Left join to othogroup data preserve all DEGs
Klac_og <- Klac_mapped %>%
  left_join(
    orthogroup2 %>% 
      filter(!is.na(klac_gene_id)) %>% 
      distinct(klac_gene_id, .keep_all = TRUE),
    by = "klac_gene_id"
  ) #

################# count results 
# keep NA as a category called "unmapped", that is singleton clusters without OG assignment by orthofinder 
count_status_keep_na <- function(df, status_col, label){
  df %>%
    mutate(
      status = fct_explicit_na(.data[[status_col]], na_level = "unmapped"),
      species = label
    ) %>%
    count(status, species, name = "n")
}

counts_og <- bind_rows(
  count_status_keep_na(Scer_og, "scer_core_status", "Scer"),
  count_status_keep_na(Cgla_og, "cgla_core_status", "Cgla"),
  count_status_keep_na(Calb_og, "calb_core_status", "Calb"),
  count_status_keep_na(Klac_og, "klac_core_status", "Klac")
) %>%
  mutate(
    status  = factor(status, levels = c("non-core","multi-core","single-core","unmapped")),
    species = factor(species, levels = c("Scer","Cgla","Calb","Klac"))
  )


####### plot data 
plot_degs_og <- ggplot(counts_og, aes(x = status, y = n, fill = species)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = n),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3, fontface = "bold") +
  labs(x = NULL, y = "Number of DEGs", title = "DEGs by core status across species") +
  theme_minimal() + bold_theme

plot_degs_og

########### # convert counts -> percent within each species. 
counts_pct_og <- counts_og %>%
  group_by(species) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()
# plot percentages
plot_degs_pct_og <- ggplot(counts_pct_og, aes(x = status, y = pct, fill = species)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", pct)),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3, fontface = "bold") +
  labs(x = NULL, y = "Percent of genes per species", title = "Core status (percent) by species") +
  theme_minimal() + bold_theme
plot_degs_og
plot_degs_pct_og

#################### percentage of deges in each core status classificationn as percentage of ref genome 
# counts      = genome-wide counts by species × status  (denominator)
# counts_og   = DEG counts      by species × status      (numerator)
pct_tbl <- counts_og %>%
  rename(n_degs  = n) %>%
  left_join(counts %>% rename(n_genes = n),
            by = c("species","status")) %>%
  mutate(
    n_degs  = replace_na(n_degs, 0L),
    pct_of_genes = ifelse(is.na(n_genes) | n_genes == 0, NA_real_,
                          100 * n_degs / n_genes),
    status  = factor(status,  levels = c("non-core","multi-core","single-core", "unmapped")),
    species = factor(species, levels = c("Scer","Cgla","Calb","Klac", "unmapped"))
  )
# Plot: DEGs as % of genes in each class (grouped by species)
pct_tbl_og <- ggplot(pct_tbl, aes(x = status, y = pct_of_genes, fill = species)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = ifelse(is.na(pct_of_genes), "",
                               sprintf("%.1f%%", pct_of_genes))),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3, fontface = "bold") +
  labs(x = NULL, y = "DEGs as % of genes in class",
       title = "DEG burden by core status across species") +
  theme_minimal() + bold_theme

plot_degs_og ##
pct_tbl_og
```

Goal

Build a per-species index of all genes from the reference GTFs, then check which of those genes are represented in our Orthofinder orthogroups table. This tells us which reference genes are covered by orthogroups and which are missing/unmapped
What this chunk does (snapshot)

1) Parse the reference GTF into a gene list (gene_id, gene_name).
Reads each species’ GTF.
Extracts attributes from the attribute column (e.g., gene_id, gene_name, ID, Name).
Keeps one row per gene; for Scer drops the gene- prefix so IDs match the rest of the pipeline.
Outputs a tidy data frame of all genes in the reference.

2) Explode per-species orthogroup membership.
Starts from orthogroups_cleaned.
For each species, splits the *_orthgrp_list (semicolon-separated) into one row per gene and keeps its OG.
This yields a per-gene ↔ OG lookup (e.g., scer_gene_id/qng_id/orf19.*/XP_... → OG).

3) Join GTF gene list to the OG map to flag coverage.
Left-join the GTF gene table to the OG map on the appropriate ID type per species (see below).
Adds in_orthogroup = TRUE/FALSE.
Writes (a) the joined status table and (b) a list of genes present in the reference but not in any OG.
Species-specific notes (ID normalization & joins)
S. cerevisiae (Scer)
GTF IDs: SGD ORFs (Y...). Removes gene- prefix if present.
OG join key: scer_gene_id extracted from Scer_orthgrp_list.

Outputs:
06.results/Scer/20250911_scer_missing_from_og.csv (genes not in any OG)

C. glabrata (Cgla)
GTF IDs: BG2 locus IDs (e.g., GWK60_*) or CAGL0*.
Maps GTF gene IDs → QNG using your Cgla_BG2Protmapping (three-way map).
OG join key: qng_id from Cgla_orthgrp_list.
Outputs:
06.results/Cgla/20250911_Cgla_missing_from_og.csv (unmapped to OG)

C. albicans (Calb)
GTF IDs: Normalized to lower-case; prefers orf19.* where possible.
OG join key: lower-case orf19.* from Calb_orthgrp_list.
Filters the “missing” table to entries that look like orf19.xxxx to avoid non-coding/misc entries.

Outputs:
06.results/Calb/20250911_Calb_missing_from_og.csv

K. lactis (Klac)
GTF IDs: locus tags/gene IDs; protein accessions may appear in protein_id/Dbxref.
Extracts XP_... accessions and uses those as the canonical join key.
OG join key: xp_id from Klac_orthgrp_list.
Full coverage table and a filtered “protein-coding only (has XP_) but missing from OG” list.

```{r missing-genes-from-OG-Assignment, warning=FALSE, message=FALSE}
################################## get the full gene list from the ref folder 
#and create data frame with gene_id and gene_name. 
#Set difference from this and the orthogroup data set to kn ow which genes are not in the orthogroup data set 

scer_gtf <- here("04.refs", "s288cSc-R64.gtf") 
gtf <- fread(
  scer_gtf, sep = "\t", header = FALSE, data.table = FALSE,
  col.names = c("seqname","source","feature","start","end","score","strand","frame","attribute"),
  quote = "" #, comment.char = "#"
)

get_attr <- function(x, key){
  m <- str_match(x, paste0(key, " \"([^\"]+)\""))
  m[,2]
}

scer_gene_list <- gtf %>%
  transmute(
    feature,
    gene_id   = get_attr(attribute, "gene_id"),
    gene_name = coalesce(get_attr(attribute, "gene_name"),
                         get_attr(attribute, "Name"))
  ) %>%
  filter(!is.na(gene_id)) %>%
  { if (any(.$feature == "gene")) filter(., feature == "gene") else . } %>%
  distinct(gene_id, gene_name) %>%
  mutate(
    gene_id   = sub("^gene-", "", gene_id)
  )

write_csv(scer_gene_og_status, file.path(Scer_results, "20250910_Scer_gene_list.csv"))

############### list all genes in the orthgrp in Scer 
scer_og_map <- orthogroups_cleaned %>%
  select(OG, S_cerevisiae, Scer_orthgrp_list) %>%
  mutate(Scer_orthgrp_list = na_if(Scer_orthgrp_list, "")) %>%
  separate_rows(Scer_orthgrp_list, sep = ";", convert = FALSE) %>%
  transmute(
    OG,
    scer_gene_id = trimws(Scer_orthgrp_list)
  ) %>%
  filter(!is.na(scer_gene_id), scer_gene_id != "") %>%
  distinct()
########2) Join the GTF gene list to the OG map to flag coverage
scer_gene_og_status <- scer_gene_list %>%
  select(gene_id, gene_name) %>%
  left_join(scer_og_map, by = c("gene_id" = "scer_gene_id")) %>%
  mutate(in_orthogroup = !is.na(OG))
write_csv(scer_gene_og_status, file.path(Scer_results, "20250911_orthgrps_gene_list.csv"))

# Genes present in GTF but missing from orthogroups
scer_missing_from_og <- scer_gene_og_status %>% filter(!in_orthogroup)
write_csv(scer_missing_from_og, file.path(Scer_results, "20250911_scer_missing_from_og.csv"))


############### C.glabrata 
# --- Inputs ---
cgla_gtf <- here("04.refs", "BG2Cg.gtf")   
get_attr <- function(x, key){
  m <- stringr::str_match(x, paste0(key, " \"([^\"]+)\""))
  m[,2]
}

# 1) Parse GTF -> raw gene table (gene_id, gene_name)
gtf_cg <- fread(
  cgla_gtf, sep = "\t", header = FALSE, data.table = FALSE,
  col.names = c("seqname","source","feature","start","end","score","strand","frame","attribute"),
  quote = "" #comment.char = "#"
)

cgla_gene_list_raw <- gtf_cg %>%
  transmute(
    feature,
    gene_id   = get_attr(attribute, "gene_id"),   # e.g., GWK60_* or CAGL0*
    gene_name = coalesce(get_attr(attribute, "gene_name"),
                         get_attr(attribute, "Name"))
  ) %>%
  filter(!is.na(gene_id)) %>%
  { if (any(.$feature == "gene")) filter(., feature == "gene") else . } %>%
  distinct(gene_id, gene_name) %>%
  mutate(gene_id = sub("^gene-", "", gene_id)) 

############ Map GTF gene IDs to QNG using your mapping table
cgla_map_keys <- Cgla_BG2Protmapping %>%
  select(gene_id, cagl_id, qng_id, gene_name_map = gene_name) %>%
  distinct()

cgla_gene_list_qng <- cgla_gene_list_raw %>%
  left_join(cgla_map_keys, by = c("gene_id" = "gene_id"))

## Explode orthogroup QNG lists -> per-gene OG map
cgla_og_map <- orthogroups_cleaned %>%
  select(OG, C_glabrata, Cgla_orthgrp_list) %>%
  mutate(Cgla_orthgrp_list = na_if(Cgla_orthgrp_list, "")) %>%
  separate_rows(Cgla_orthgrp_list, sep = ";", convert = FALSE) %>%
  transmute(
    OG,
    qng_id = trimws(Cgla_orthgrp_list)
  ) %>%
  filter(!is.na(qng_id), qng_id != "") %>%
  distinct()

## Coverage: join GTF→QNG to OG map
cgla_gene_og_status <- cgla_gene_list_qng %>%
  left_join(cgla_og_map, by = "qng_id") %>%
  mutate(in_orthogroup = !is.na(OG)) %>%
  select(OG, gene_id, qng_id, cagl_id, in_orthogroup)

### Diagnostics
cgla_missing_from_og <- cgla_gene_og_status %>% filter(is.na(OG) | !in_orthogroup)
write_csv(cgla_missing_from_og, file.path(Cgla_results, "20250911_Cgla_missing_from_og.csv"))

########## C.albicans 

calb_gtf <- here("04.refs", "SC5314Ca.gtf")


get_attr <- function(x, key){
  m <- stringr::str_match(x, paste0(key, " \"([^\"]+)\""))
  m[,2]
}

# 1) Parse GTF -> Calb gene table (prefer orf19.* IDs)
gtf_ca <- fread(
  calb_gtf, sep = "\t", header = FALSE, data.table = FALSE,
  col.names = c("seqname","source","feature","start","end","score","strand","frame","attribute"),
  quote = "" #comment.char = "#"
)

calb_gene_list <- gtf_ca %>%
  transmute(
    feature,
    # Try common keys; fall back to Name if needed
    raw_id    = coalesce(get_attr(attribute, "gene_id"),
                         get_attr(attribute, "ID"),
                         get_attr(attribute, "locus_tag"),
                         get_attr(attribute, "Name")),
    gene_name = coalesce(get_attr(attribute, "gene_name"),
                         get_attr(attribute, "Name"))
  ) %>%
  filter(!is.na(raw_id)) %>%
  { if (any(.$feature == "gene")) filter(., feature == "gene") else . } %>%
  transmute(
    gene_id   = tolower(sub("^gene-", "", raw_id)),  # normalize: drop "gene-" and lowercase
    gene_name = gene_name
  ) %>%
  distinct(gene_id, gene_name)

# 2) Explode orthogroup Calb list -> per-gene OG map (expects orf19.*)
calb_og_map <- orthogroups_cleaned %>%
  select(OG, C_albicans, Calb_orthgrp_list) %>%
  mutate(Calb_orthgrp_list = na_if(Calb_orthgrp_list, "")) %>%
  separate_rows(Calb_orthgrp_list, sep = ";", convert = FALSE) %>%
  transmute(
    OG,
    calb_gene_id = tolower(trimws(Calb_orthgrp_list))
  ) %>%
  filter(!is.na(calb_gene_id), calb_gene_id != "") %>%
  distinct()

# 3) Coverage: join GTF genes to OG map
calb_gene_og_status <- calb_gene_list %>%
  left_join(calb_og_map, by = c("gene_id" = "calb_gene_id")) %>%
  mutate(in_orthogroup = !is.na(OG))

# 4) Diagnostics
calb_orf_pat <- regex("^orf\\d+\\.\\d+$", ignore_case = TRUE)
calb_missing_from_og <- calb_gene_og_status %>%
  dplyr::filter(!in_orthogroup) %>%
  dplyr::filter(str_detect(gene_id, calb_orf_pat))
###########
write_csv(calb_missing_from_og, file.path(Calb_results, "20250911_Calb_missing_from_og.csv"))


############ K. lactis 

klac_gtf <- here("04.refs", "CBS2105_kl.gtf")  

# Helper to pull attributes from GTF 'attribute' column
get_attr <- function(x, key){
  m <- stringr::str_match(x, paste0(key, " \"([^\"]+)\""))
  m[,2]
}

xp_pat <- "XP_\\d+(?:\\.\\d+)?"

# 1) Parse GTF -> Klac gene table (map gene_id -> XP_ protein if available)
gtf_kl <- fread(
  klac_gtf, sep = "\t", header = FALSE, data.table = FALSE,
  col.names = c("seqname","source","feature","start","end","score","strand","frame","attribute"),
  quote = "" #, comment.char = "#"
)

klac_gene_list <- gtf_kl %>%
  transmute(
    feature,
    gene_id    = coalesce(get_attr(attribute, "gene_id"),
                          get_attr(attribute, "ID"),
                          get_attr(attribute, "locus_tag"),
                          get_attr(attribute, "Name")),
    gene_name  = coalesce(get_attr(attribute, "gene_name"),
                          get_attr(attribute, "Name")),
    # XP_ may appear in protein_id or Dbxref fields; fall back to "any XP_ in attribute"
    xp_from_protein = get_attr(attribute, "protein_id"),
    xp_from_dbxref  = {
      m <- str_extract(attribute, xp_pat); m
    }
  ) %>%
  filter(!is.na(gene_id)) %>%
  { if (any(.$feature == "gene")) filter(., feature == "gene") else . } %>%
  mutate(
    gene_id = sub("^gene-", "", gene_id),
    xp_id   = coalesce(xp_from_protein, xp_from_dbxref)
  ) %>%
  distinct(gene_id, gene_name, xp_id)

# 2) Explode orthogroup Klac list (XP_) -> per-gene OG map
klac_og_map <- orthogroups_cleaned %>%
  select(OG, K_lactis, Klac_orthgrp_list) %>%
  mutate(Klac_orthgrp_list = na_if(Klac_orthgrp_list, "")) %>%
  separate_rows(Klac_orthgrp_list, sep = ";", convert = FALSE) %>%
  transmute(
    OG,
    xp_id = trimws(Klac_orthgrp_list)
  ) %>%
  filter(!is.na(xp_id), xp_id != "") %>%
  distinct()

# 3) Coverage: join GTF genes to OG map via XP_
klac_gene_og_status <- klac_gene_list %>%
  left_join(klac_og_map, by = "xp_id") %>%
  mutate(in_orthogroup = !is.na(OG))

# 4) Diagnostics
# (a) All genes missing from OG (includes non-protein-coding if they lack XP_)
klac_missing_from_og <- klac_gene_og_status %>%
  filter(!in_orthogroup)

# (b) Protein-coding only (must have XP_) and still missing from OG
klac_missing_from_og_xpOnly <- klac_gene_og_status %>%
  filter(!in_orthogroup, !is.na(xp_id))



```


```{r doughty-fig3-d, warning=FALSE, message=FALSE}
# Scer_DE_topGo.csv = full DESeq2 results for S. cerevisiae (all tested genes, not just DEGs), with columns like gene_id, log2FoldChange, padj.
# 
# orthogroup2 = your master orthology table containing for Scer the columns Scer_orthgrp_list (semi-colon list of SGD ORFs in the orthogroup) and scer_core_status (non-core / multi-core / single-core).
# 
# Scer_og = the already-subset Scer DEG table you created earlier (DEGs joined to orthogroup2, so it includes scer_core_st


# S. cerevisiae — FULL DESeq results (not subsetted)
scer_full <- read_csv(here("06.results", "Scer", "Scer_DE_topGo.csv"), 
                      show_col_types = FALSE) %>%
  mutate(scer_gene_id = gene_id)

#Set thresholds

fdr_cutoff <- 0.01   # instead of alpha
lfc_thr    <- 1
use_abs    <- FALSE

# --- Build per-gene maps directly from orthogroup2 (explode *_orthgrp_list) --- Build a per-gene mapping from orthogroups to core status
scer_map <- orthogroup2 %>%
  transmute(scer_core_status,
            scer_gene_id = strsplit(replace_na(Scer_orthgrp_list, ""), ";")) %>%
  unnest_longer(scer_gene_id) %>%
  mutate(scer_gene_id = na_if(trimws(scer_gene_id), "")) %>%
  filter(!is.na(scer_gene_id), !is.na(scer_core_status)) %>%
  distinct(scer_gene_id, scer_core_status) 

## Join full DESeq table to core status and coerce numeric
full_status_scer <- scer_full %>%
  left_join(scer_map, by = "scer_gene_id") %>%
  mutate(
    padj_num = suppressWarnings(as.double(padj)),
    l2fc_num = suppressWarnings(as.double(log2FoldChange))
  ) %>%
  filter(!is.na(scer_core_status), !is.na(padj_num))   # "detected mRNAs"

## Define DE flags and overall DE rate
is_deg_scer <- (full_status_scer$padj_num < fdr_cutoff) &
  (if (use_abs) abs(full_status_scer$l2fc_num) > lfc_thr else full_status_scer$l2fc_num > lfc_thr)

overall_deg  <- sum(is_deg_scer, na.rm = TRUE)
overall_n    <- nrow(full_status_scer)
overall_rate <- overall_deg / overall_n

## --- Denominators: detected per class (same as before) ---
denoms_scer <- full_status_scer %>%
  count(status = scer_core_status, name = "n_class")

## --- Numerators: DE within the SAME universe + SAME rule ---
degs_by_class_scer <- full_status_scer %>%
  mutate(is_deg = (padj_num < fdr_cutoff) &
                   (if (use_abs) abs(l2fc_num) > lfc_thr else l2fc_num > lfc_thr)) %>%
  group_by(status = scer_core_status) %>%
  summarise(n_deg = sum(is_deg, na.rm = TRUE), .groups = "drop")

## --- Enrichment table (Fig. 1c metrics) ---
fig1c_scer <- denoms_scer %>%
  left_join(degs_by_class_scer, by = "status") %>%
  mutate(
    n_deg        = replace_na(n_deg, 0L),
    n_notdeg     = n_class - n_deg,
    n_out_deg    = overall_deg - n_deg,
    n_out_notdeg = (overall_n - n_class) - n_out_deg,
    rate_class     = n_deg / n_class,
    fold_enrichment = rate_class / overall_rate,
    p_value = purrr::pmap_dbl(
      list(n_deg, n_notdeg, n_out_deg, n_out_notdeg),
      ~ fisher.test(matrix(c(..1, ..2, ..3, ..4), nrow = 2))$p.value
    ),
    species = "Scer"
  )

########################
cgla_full <- read_csv(here("06.results", "Cgla", "Cgla_DE_topGo.csv"), 
                      show_col_types = FALSE) %>%
  mutate(cgla_gene_id = gene_id) 

cgla_map <- orthogroup2 %>%
  transmute(cgla_core_status,
            cgla_gene_id = strsplit(replace_na(Cgla_orthgrp_list, ""), ";")) %>%
  unnest_longer(cgla_gene_id) %>%
  mutate(cgla_gene_id = na_if(trimws(cgla_gene_id), "")) %>%
  filter(!is.na(cgla_gene_id), !is.na(cgla_core_status)) %>%
  distinct(cgla_gene_id, cgla_core_status) %>% 
  rename(qng_id = cgla_gene_id) %>%
  left_join(Cgla_BG2Protmapping, by = "qng_id") %>%
  select(gene_id, cgla_core_status) %>%
  rename(cgla_gene_id = gene_id)
  

## --- Join full DESeq table to core status and coerce numeric ---
full_status_cgla <- cgla_full %>%
  left_join(cgla_map, by = "cgla_gene_id") %>%
  mutate(
    padj_num = suppressWarnings(as.double(padj)),
    l2fc_num = suppressWarnings(as.double(log2FoldChange))
  ) %>%
  filter(!is.na(cgla_core_status), !is.na(padj_num))            # detected mRNAs in OGs

## --- Define DE flags and overall DE rate (use the same fdr_cutoff/lfc_thr/use_abs you set) ---
is_deg_cgla <- (full_status_cgla$padj_num < fdr_cutoff) &
  (if (use_abs) abs(full_status_cgla$l2fc_num) > lfc_thr else full_status_cgla$l2fc_num > lfc_thr)

overall_deg_cgla  <- sum(is_deg_cgla, na.rm = TRUE)
overall_n_cgla    <- nrow(full_status_cgla)
overall_rate_cgla <- overall_deg_cgla / overall_n_cgla

## --- Denominators (detected per class) ---
denoms_cgla <- full_status_cgla %>%
  count(status = cgla_core_status, name = "n_class")

## --- Numerators (your already-joined DEGs table for Cgla) ---
degs_by_class_cgla <- full_status_cgla %>%
  mutate(is_deg = (padj_num < fdr_cutoff) &
                  (if (use_abs) abs(l2fc_num) > lfc_thr else l2fc_num > lfc_thr)) %>%
  group_by(status = cgla_core_status) %>%
  summarise(n_deg = sum(is_deg, na.rm = TRUE), .groups = "drop")
## --- Enrichment table (Fig. 1c metrics) ---
fig1c_cgla <- denoms_cgla %>%
  left_join(degs_by_class_cgla, by = "status") %>%
  mutate(
    n_deg        = replace_na(n_deg, 0L),
    n_notdeg     = n_class - n_deg,
    n_out_deg    = overall_deg_cgla - n_deg,
    n_out_notdeg = (overall_n_cgla - n_class) - n_out_deg,
    rate_class     = n_deg / n_class,
    fold_enrichment = rate_class / overall_rate_cgla,
    p_value = purrr::pmap_dbl(
      list(n_deg, n_notdeg, n_out_deg, n_out_notdeg),
      ~ fisher.test(matrix(c(..1, ..2, ..3, ..4), nrow = 2))$p.value
    ),
    species = "Cgla"
  )

#############################
########### 1 Full DESeq results for Calb (ALL tested genes)
calb_full <- read_csv(here("06.results", "Calb", "Calb_DE_topGo.csv"), 
                      show_col_types = FALSE) %>%
  mutate(calb_gene_id = gene_id) 

############# 2 Per-gene core status map from orthogroup2 (explode Calb_orthgrp_list)
calb_map <- orthogroup2 %>%
  transmute(calb_core_status,
            calb_gene_id = strsplit(replace_na(Calb_orthgrp_list, ""), ";")) %>%
  unnest_longer(calb_gene_id) %>%
  mutate(calb_gene_id = na_if(trimws(calb_gene_id), "")) %>%
  filter(!is.na(calb_gene_id), !is.na(calb_core_status)) %>%
  distinct(calb_gene_id, calb_core_status)

# 3) Join full DESeq to core status; coerce numerics; keep detected & mapped
full_status_calb <- calb_full %>%
  left_join(calb_map, by = "calb_gene_id") %>%
  mutate(
    padj_num = suppressWarnings(as.double(padj)),
    l2fc_num = suppressWarnings(as.double(log2FoldChange))
  ) %>%
  filter(!is.na(calb_core_status), !is.na(padj_num))   # detected mRNAs inside OGs

# 4) Overall DE flags/rate (same rule as you use for plotting)
is_deg_calb <- (full_status_calb$padj_num < fdr_cutoff) &
  (if (use_abs) abs(full_status_calb$l2fc_num) > lfc_thr else full_status_calb$l2fc_num > lfc_thr)

overall_deg_calb  <- sum(is_deg_calb, na.rm = TRUE)
overall_n_calb    <- nrow(full_status_calb)
overall_rate_calb <- overall_deg_calb / overall_n_calb

# 5) Denominators per class (detected)
denoms_calb <- full_status_calb %>%
  count(status = calb_core_status, name = "n_class")

# 6) Numerators per class (DE within the SAME universe/rule)
degs_by_class_calb <- full_status_calb %>%
  mutate(is_deg = (padj_num < fdr_cutoff) &
                   (if (use_abs) abs(l2fc_num) > lfc_thr else l2fc_num > lfc_thr)) %>%
  group_by(status = calb_core_status) %>%
  summarise(n_deg = sum(is_deg, na.rm = TRUE), .groups = "drop")

# 7) Fig-1c style enrichment table (Calb)
fig1c_calb <- denoms_calb %>%
  left_join(degs_by_class_calb, by = "status") %>%
  mutate(
    n_deg        = replace_na(n_deg, 0L),
    n_notdeg     = n_class - n_deg,
    n_out_deg    = overall_deg_calb - n_deg,
    n_out_notdeg = (overall_n_calb - n_class) - n_out_deg,
    rate_class     = n_deg / n_class,
    fold_enrichment = rate_class / overall_rate_calb,
    p_value = pmap_dbl(
      list(n_deg, n_notdeg, n_out_deg, n_out_notdeg),
      ~ fisher.test(matrix(c(..1, ..2, ..3, ..4), nrow = 2))$p.value
    ),
    species = "Calb"
  )
#####################
########### 1 Full DESeq results for Calb (ALL tested genes)
klac_full <- read_csv(here("06.results", "Klac", "Klac_DE_topGo.csv"), 
                      show_col_types = FALSE) %>%
  mutate(klac_gene_id = gene_id) 

############# 2 Per-gene core status map from orthogroup2 (explode Calb_orthgrp_list
klac_map <- orthogroup2 %>%
  transmute(klac_core_status,
            klac_gene_id = strsplit(replace_na(Klac_orthgrp_list, ""), ";")) %>%
  unnest_longer(klac_gene_id) %>%
  mutate(klac_gene_id = na_if(trimws(klac_gene_id), "")) %>%
  filter(!is.na(klac_gene_id), !is.na(klac_core_status)) %>%
  distinct(klac_gene_id, klac_core_status) %>% 
  rename(xp_id = klac_gene_id) %>%
  left_join(Klac_xpklla0tmapping, by = "xp_id") %>%
  select(locus_tag, klac_core_status) %>%
  rename(klac_gene_id = locus_tag)
  
# 3) Join, coerce numerics, keep detected & mapped (denominator universe)
full_status_klac <- klac_full %>%
  left_join(klac_map, by = "klac_gene_id") %>%
  mutate(
    padj_num = suppressWarnings(as.double(padj)),
    l2fc_num = suppressWarnings(as.double(log2FoldChange))
  ) %>%
  filter(!is.na(klac_core_status), !is.na(padj_num))

# 4) Overall DE flags/rate (same rule you use everywhere)
is_deg_klac <- (full_status_klac$padj_num < fdr_cutoff) &
  (if (use_abs) abs(full_status_klac$l2fc_num) > lfc_thr else full_status_klac$l2fc_num > lfc_thr)

overall_deg_klac  <- sum(is_deg_klac, na.rm = TRUE)
overall_n_klac    <- nrow(full_status_klac)
overall_rate_klac <- overall_deg_klac / overall_n_klac

# 5) Denominators per class
denoms_klac <- full_status_klac %>%
  count(status = klac_core_status, name = "n_class")

# 6) Numerators per class (DE within SAME universe/rule)
degs_by_class_klac <- full_status_klac %>%
  mutate(is_deg = (padj_num < fdr_cutoff) &
                   (if (use_abs) abs(l2fc_num) > lfc_thr else l2fc_num > lfc_thr)) %>%
  group_by(status = klac_core_status) %>%
  summarise(n_deg = sum(is_deg, na.rm = TRUE), .groups = "drop")

# 7) Fig-1c style enrichment table
fig1c_klac <- denoms_klac %>%
  left_join(degs_by_class_klac, by = "status") %>%
  mutate(
    n_deg        = replace_na(n_deg, 0L),
    n_notdeg     = n_class - n_deg,
    n_out_deg    = overall_deg_klac - n_deg,
    n_out_notdeg = (overall_n_klac - n_class) - n_out_deg,
    rate_class     = n_deg / n_class,
    fold_enrichment = rate_class / overall_rate_klac,
    p_value = pmap_dbl(
      list(n_deg, n_notdeg, n_out_deg, n_out_notdeg),
      ~ fisher.test(matrix(c(..1, ..2, ..3, ..4), nrow = 2))$p.value
    ),
    species = "Klac"
  )

# Combine and plot
fig1c_all <- bind_rows(fig1c_scer, fig1c_cgla, fig1c_calb, fig1c_klac) %>%
  mutate(
    status  = factor(status, levels = c("non-core","multi-core","single-core")),
    species = factor(species, levels = c("Scer","Cgla","Calb","Klac")),
    sig = case_when(p_value < 0.001 ~ "***",
                    p_value < 0.01  ~ "**",
                    p_value < 0.05  ~ "*",
                    TRUE            ~ "")
  )

p_fig1c_all <- ggplot(fig1c_all, aes(x=status, y=fold_enrichment, fill=species)) +
  geom_hline(yintercept=1, linewidth=0.6, linetype=2) +
  geom_col(position=position_dodge(0.8), width=0.7) +
  geom_text(aes(label=sig),
            position=position_dodge(0.8), vjust=-0.6, size=6, fontface="bold") +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.15))) +
  coord_cartesian(clip = "off") +
  labs(x=NULL, y="Fold enrichment of DE rate (class / overall)",
       title="Stress adaptation responsive genes in -Pi") +
  theme_minimal() + bold_theme

p_fig1c_all

```

